---
title: Standardize the Seurat Object
author: Tianze Cao
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
#set to project dir
knitr::opts_knit$set(root.dir = "/home/ti4104ca/Projects/Mine")
knitr::opts_chunk$set(echo = TRUE, results = 'hold', fig.show = 'hold', warning = FALSE)
```

# Load
## Libraries
```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(ggplot2)
  library(ggridges)
  library(dplyr)
  library(magrittr)
  library(kableExtra)
  source("code/scripts/SexAssignment.R")
  library(org.Hs.eg.db)
  library(patchwork)
  library(wesanderson)
})
```

## Set input file
```{r}
input <- "data/raw_data/guilliams_liver_2022.rds"
raw <- readRDS(input)
```

# Standarize seurat object
```{r}
if (grepl("disco", input)) {
  #restore data slot
  raw@assays$RNA@data <- raw@assays$RNA@counts
}

if (grepl("caq", input)) {
  raw@assays$RNA <- raw@assays$originalexp
  colnames(raw@meta.data)[colnames(raw@meta.data) == 'cell_type'] <- 'cell_type_unhormonized'
}

#if the object is seurat v5
raw@meta.data <- raw@meta.data %>%
  mutate_all(as.character)
```

## Set preparation is needed
```{r eval=TRUE}
#if there is not nFeature, nCount
raw@meta.data$nCount_RNA = colSums(x = raw, slot = "counts")
raw@meta.data$nFeature_RNA = colSums(x = GetAssayData(object = raw, slot = "counts") > 0)

#if have to convert from Ensembl gene id to Symbol
#TO DO: converted mitochondrial genes lost hyphen.
#gene_id <- select(org.Hs.eg.db, keys = rownames(raw), keytype = 'ENSEMBL', columns = 'SYMBOL') %>%
 # .[!duplicated(.$ENSEMBL), ]
#raw@assays[["RNA"]]@counts@Dimnames[[1]] <- ifelse(is.na(gene_id$SYMBOL), gene_id$ENSEMBL, gene_id$SYMBOL)
#raw@assays[["RNA"]]@data@Dimnames[[1]] <- ifelse(is.na(gene_id$SYMBOL), gene_id$ENSEMBL, gene_id$SYMBOL)

#if somewhere has Symbol
raw@assays[["RNA"]]@counts@Dimnames[[1]] <- raw@assays[["RNA"]]@meta.features[["feature_name"]]
raw@assays[["RNA"]]@data@Dimnames[[1]] <- raw@assays[["RNA"]]@meta.features[["feature_name"]]

#if there is not percent.mt
raw[["percent.mt"]] = PercentageFeatureSet(raw, pattern = "^MT-")
```

## Uniform attribute names
```{r}
colnames(raw@meta.data)
```

### Set the names in metadata
```{r}
colnames(raw@meta.data)[colnames(raw@meta.data) == 'development_stage'] <- 'age'
colnames(raw@meta.data)[colnames(raw@meta.data) == ''] <- 'sex'
raw@meta.data$sex[raw@meta.data$sex == 'female'] <- 'F'
raw@meta.data$sex[raw@meta.data$sex == 'male'] <- 'M'



colnames(raw@meta.data)[colnames(raw@meta.data) == 'nFeature_RNA'] <- 'nFeature'
colnames(raw@meta.data)[colnames(raw@meta.data) == 'nCount_RNA'] <- 'nCount'
colnames(raw@meta.data)[colnames(raw@meta.data) == ''] <- 'percent.mt'



#make sure one donor one eligible sample
colnames(raw@meta.data)[colnames(raw@meta.data) == 'donor_id'] <- 'sample_id'

colnames(raw@meta.data)[colnames(raw@meta.data) == ''] <- 'project_id'

#if have multiple cell type annotation
colnames(raw@meta.data)[colnames(raw@meta.data) == 'cell_type'] <- 'cellxgene_cell_type'

colnames(raw@meta.data)[colnames(raw@meta.data) == 'author_cell_type'] <- 'cell_type'
colnames(raw@meta.data)[colnames(raw@meta.data) == 'assay'] <- 'platform'
raw@meta.data$platform <- gsub("'", "", raw@meta.data$platform)
raw@meta.data$platform <- gsub(" ", "", raw@meta.data$platform)
```

```{r}
colnames(raw@meta.data)
```

## Set suitable condition
variables for liver: age, obesity, virus, alcohol consumption, sampling site, region.
```{r}
#set the condition
con <- c(quo(typeSample), quo(reported_diseases), quo(diet))

if ('project_id' %in% colnames(raw@meta.data)) {
  for (i in seq_along(con)) {
    raw@meta.data %>%
      group_by(!!con[[i]]) %>%
      summarise(
        count = n_distinct(sample_id),
        project_id = toString(unique(project_id)),
        sample_id = toString(unique(sample_id))
      ) %>%
      print()
  }
} else {
  for (i in seq_along(con)) {
    raw@meta.data %>%
      group_by(!!con[[i]]) %>%
      summarise(
        count = n_distinct(sample_id),
        sample_id = toString(unique(sample_id))
      ) %>%
      print()
  }
}
```

### Set the state of condition
```{r}
se <- raw

#select cells based on their state of condition
se <- subset(se, typeSample == "scRnaSeq")
```

```{r}
#only keep projects with at least 6 samples
if ('project_id' %in% colnames(se@meta.data)) {
  proj_df <- se@meta.data %>%
    group_by(project_id) %>%
    reframe('count' = n_distinct(sample_id), 'sample_id' = toString(unique(sample_id)))
  projs <- proj_df$project_id[proj_df$count >= 6]
  se <- subset(se, subset = project_id %in% projs)
  proj_df
}
```

### Set projects after curation
Project manual curation:
x - Disco liver:
  x - E-MTAB-7407 (fetal)
  x - GSE156625 (after removing fetal samples, M:F = 7:2)
x - caq liver:
  x - E-MTAB-11536, 1b9d8702-5af8-4142-85ed-020eb06ec4f6 --> https://cellxgene.cziscience.com/collections/62ef75e4-cbea-454e-a0ce-998ec40223d3 (M:F = 5:2)
  x- GSE185477, ddb22b3d-a75c-4dd1-9730-dff7fc8ca530 --> https://cellxgene.cziscience.com/collections/44531dd9-1388-4416-a117-af0a99de2294 (4 donors)
cellxgene liver:
  https://www.sciencedirect.com/science/article/pii/S0092867421014811?via%3Dihub --> adjacent healthy tissue, table s6 has live histology assessed --> donors pass assessment: H02, H06, H07, H10, H16, H22.
```{r}
if ('project_id' %in% colnames(se@meta.data)) {
  se <- subset(se, subset = project_id == "1b9d8702-5af8-4142-85ed-020eb06ec4f6")
} else {
  se <- subset(se, subset = sample_id %in% c("H02", "H06", "H07", "H10", "H16", "H22"))
}
```

After condition control:
```{r}
if ('project_id' %in% colnames(se@meta.data)) {
  for (i in seq_along(con)) {
    se@meta.data %>%
      group_by(!!con[[i]]) %>%
      summarise(
        count = n_distinct(sample_id),
        project_id = toString(unique(project_id)),
        sample_id = toString(unique(sample_id))
      ) %>%
      print()
  }
} else {
  for (i in seq_along(con)) {
    se@meta.data %>%
      group_by(!!con[[i]]) %>%
      summarise(
        count = n_distinct(sample_id),
        sample_id = toString(unique(sample_id))
      ) %>%
      print()
  }
}
```

## Customize cell types
```{r}
kbl(t(table(se@meta.data$cell_type))) %>%
  kable_paper() %>%
  scroll_box(width = "620px", height = "150px")
```

### Set simplified cell types
```{r eval=FALSE}
#disco_lung
myeloid_cell <- c("G2/M phase myeloid cell", "S phase myeloid cell")

macrophage <- c("M1 macrophage", "Alveolar macrophage", "LYVE1 macrophage")

dendritic_cell <- c("mregDC", "cDC1", "cDC2")

monocyte <- c("CD14 monocyte", "CD16 monocyte")

b_cell <- c("Naive B cell", "Memory B cell", "Plasma cell")

mast_cell <- c("Mast cell")

nk_cell <- c("CD16 NK cell")

t_cell <- c("Memory CD4 T cell", "Naive CD4 T cell", "GZMB CD8 T cell", "GZMK CD8 T cell", "Treg cell")

t_nk_cell <- "Cycling T/NK cell"

non_immune_cell <- c("Aerocyte", "Airway basal cell", "Arterial EC", "AT1", "AT2", "Capillary EC", "CDC20B+ Multiciliated cell", "Club cell", "CXCL14+POSTN+ fibroblast", "Cycling AT2", "FBLN2+APOD+ fibroblast", "Goblet cell", "G0S2+PPP1R14A+ fibroblast", "IGFBP6+APOD+ fibroblast", "Lymphatic EC", "Multiciliated cell", "Myofibroblast", "Neuroendocrine cell", "Neuron", "Pericyte", "PLA2G2A+IGFBP4+APOC1+ fibroblast", "Red blood cell", "Vascular smooth muscle cell", "Venous EC")
```

```{r eval=FALSE}
#disco_liver

# Define non-immune cell types
non_immune_cell <- c(
  "CXCL6+ cholangiocyte",
  "Arterial EC",
  "TFF+ cholangiocyte",
  "SPINK1+SGMS2+ cholangiocyte",
  "Vascular smooth muscle cell",
  "Venous EC",
  "Hepatocyte",
  "Lymphatic EC",
  "Capillary EC",
  "Pericyte",
  "Fibroblast",
  "Glial cell"
)

# Define progenitor cell types
progenitor_cell <- c(
  "Common myeloid progenitor",
  "Common lymphoid progenitor",
  "Granulocyte-monocyte progenitor",
  "Megakaryocyte progenitor",
  "AFP+ fetal hepatocyte",
  "Late hemoglobin+ erythroblast",
  "Intermediate EPCAM+ erythroblast"
)

# Define myeloid cell types
myeloid_cell <- c(
)

# Define macrophage cell types
macrophage <- c(
  "LYVE1 macrophage",
  "M1 macrophage",
  "Cycling LYVE1 macrophage"
)

# Define dendritic cell types
dendritic_cell <- c(
  "Lymphoid pre-pDC",
  "pDC",
  "Cycling cDC2",
  "Myeloid pre-pDC",
  "cDC1",
  "cDC2"
)

# Define monocyte cell types
monocyte <- c(
  "MHCII low CD14 monocyte",
  "MHCII high CD14 monocyte",
  "CD16 monocyte"
)

# Define B cell types
b_cell <- c(
  "Naive B cell",
  "ProB cell",
  "Cycling preB cell",
  "PreB cell",
  "Plasma cell"
)

# Define mast cell type
mast_cell <- c(
  "Mast cell"
)

# Define NK cell types
nk_cell <- c(
  "CD16 NK cell",
  "CD56 NK cell"
)

# Define T cell types
t_cell <- c(
  "MAIT cell",
  "Memory CD4 T cell",
  "Naive CD8 T cell",
  "Naive CD4 T cell",
  "GZMB CD8 T cell",
  "GZMK+IL7R+ CD8 T cell",
  "GZMK+IL7R- CD8 T cell",
  "Treg cell"
)

# Define T/NK cell type
t_nk_cell <- "Cycling T/NK cell"

# Define ILC (Innate Lymphoid Cell) type
ilc <- "ILC"
```

```{r eval=TRUE}
#guilliams_liver

neutrophil <- c("Neutrophils")

basophil <- c("Basophils")

monocyte <- c("Mono+mono derived cells")

macrophage <- c("Macrophages")

dendritic_cell <- c("Mig.cDCs", "cDC1s", "cDC2s")

plasma_dc <- c("pDCs")

b_cell <- c("B cells", "Plasma cells")

t_cell <- c("T cells")

t_nk_cell <- "Circulating NK/NKT"

nk_cell <- c("Resident NK")

non_immune_cell <- c("Hepatocytes", "Cholangiocytes", "Endothelial cells", "Fibroblasts")
```

```{r}
#add simplified cell type to metadata
simp_cell_type <- ifelse(se$cell_type %in% neutrophil, "Neutrophil",
                         ifelse(se$cell_type %in% basophil, "Basophil",
                            ifelse(se$cell_type %in% monocyte, "Monocyte",
                              ifelse(se$cell_type %in% macrophage, "Macrophage",
                                ifelse(se$cell_type %in% dendritic_cell, "Dendritic cell",
                                  ifelse(se$cell_type %in% plasma_dc, "Plasma dendritic cell",
                                    ifelse(se$cell_type %in% b_cell, "B cell",
                                      ifelse(se$cell_type %in% t_cell, "T cell",
                                        ifelse(se$cell_type %in% t_nk_cell, "Cycling T/NK cell",
                                          ifelse(se$cell_type %in% nk_cell, "NK cell",
                                            ifelse(se$cell_type %in% non_immune_cell, "Non-immune cell","NA")))))))))))

se@meta.data$simp_cell_type <- simp_cell_type
```

```{r}
kbl(t(table(se@meta.data$simp_cell_type))) %>%
  kable_paper() %>%
  scroll_box(width = "620px", height = "120px")
```

#### Set order and colors
```{r}
simp_ct_order <- c("NK cell", "Cycling T/NK cell", "T cell", "B cell", "Plasma dendritic cell", "Dendritic cell", "Macrophage", "Monocyte", "Basophil", "Neutrophil")

gradient_palette <- colorRampPalette(c("#5A5FFA", "#E442A6", "orange", "#3B7A2B"))

simp_gradient_colors <- gradient_palette(length(simp_ct_order))

simp_ct_colors <- setNames(simp_gradient_colors, simp_ct_order)

simp_ct_colors["Non-immune cell"] <- "grey"
```

```{r}
ct_order <- c(nk_cell, t_nk_cell, t_cell, b_cell, plasma_dc, dendritic_cell, macrophage, monocyte, basophil, neutrophil)

gradient_palette <- colorRampPalette(c("#5A5FFA", "#E442A6", "orange", "#3B7A2B")) 

gradient_colors <- gradient_palette(length(ct_order))

ct_colors <- setNames(gradient_colors, ct_order)

ct_colors[non_immune_cell] <- "grey"
```

```{r message=FALSE}
if (!("umap" %in% names(se@reductions))){
  se <- SCTransform(se, assay = "RNA", verbose = FALSE)
  se = RunPCA(se, verbose = FALSE)
  se = FindNeighbors(se, dims = 1:50, verbose = FALSE)
  se = FindClusters(se, resolution = 0.5, algorithm = 2, verbose = FALSE)
  se = RunUMAP(se, dims = 1:50, verbose = FALSE)
}
```

```{r fig.width=8, fig.height=9}
DimPlot(se, reduction = "umap", group.by = "cell_type", label = TRUE, raster = FALSE) +
  scale_color_manual(values = ct_colors) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 10))
```

```{r fig.width=8, fig.height=9}
DimPlot(se, reduction = "umap", group.by = "simp_cell_type", label = TRUE, raster = FALSE) +
  scale_color_manual(values = simp_ct_colors) +
  theme(legend.position = "bottom")
```

```{r}
#remove non-immune cells
se <- subset(se, subset = simp_cell_type != "Non-immune cell")
```

# Describe available samples
## Sexes
```{r}
se@meta.data %>%
  group_by(sex) %>%
  reframe('count' = n_distinct(sample_id), 'sample_id' = toString(unique(sample_id))) %>%
  kbl() %>%
  kable_paper() %>%
  scroll_box(width = "620px", height = "300px")
```

### Sex assignment
```{r message=FALSE}
load("data/processed_data/guilliams_liver/sexAssign_output_20240513.RData")
#sex_assign_out <- SexAssign(data = se,
 #                           sex_col = "sex",
  #                          sample_col = "sample_id",
   #                         do.report = TRUE,
    #                        assay = "RNA",
     #                       min.percent = 0.7, # minimum percentage of cells
      #                      min.ratio = 2 # mininum ratio between the sexes
#)
```

```{r}
prev_sex <- se@meta.data %>%
        distinct(sample_id, sex) %>%
        dplyr::count(sex) %>%
        mutate(var = "Provided\nsex annotation") %>%
        dplyr::rename(predicted_sex = sex) %>%
        mutate(predicted_sex = ifelse(predicted_sex == "", "Not provided", predicted_sex))
pred_sex <- sex_assign_out[[1]]@meta.data %>%
        distinct(sample_id, predicted_sex) %>%
        dplyr::count(predicted_sex) %>%
        mutate(var="SexAssign\nprediction")
sex_annot <- bind_rows(prev_sex, pred_sex)
sex_annot_bar <- sex_annot %>%
        ggplot( aes(fill=factor(predicted_sex, c("F", "Not provided", "inconclusive", "M")), y=n, x=var, label = n)) +
        geom_bar(position="fill", stat="identity") +
        geom_text(size = 8, position = position_fill(vjust = 0.5)) +
        # change text
        labs(x = "",
             y = "Proportion of samples",
             # title = "Sex annotation vs. SexAssign prediction",
             fill = "Sample sex") +
        theme_classic() +
        scale_fill_manual(values = c("M" = "gold", "F" = "purple", "Not provided" = "darkgray", "inconclusive" = "lightgray"),
                          labels = c( "inconclusive" = "Inconclusive", "F" = "Female", "M" = "Male"))

mismatch_barplot <- as.data.frame(sex_assign_out[[2]]) %>%
  # make match column factor
    mutate(match = factor(match, c("New_annotation", "Mismatch",  "Match" ))) %>%
  dplyr::count(match) %>%
  mutate(var="") %>%
  ggplot(aes(fill=factor(match, c("Match", "Mismatch", "New_annotation")), x=var, y=n, label = n)) +
    geom_bar(position="fill", stat="identity") +
    geom_text(size = 8, position = position_fill(vjust = 0.5)) +
    # change text
    labs(x="",
         y="Proportion of samples",
         # title = "Mismatches in sample sex",
         fill = "") +
    theme_classic() +
    scale_fill_manual(values = c("Match" = "#90ee90", "Mismatch" = "#ff0033", "New_annotation" = "#4169E1"),
                      labels = c("New_annotation" = "New annotation"))
```

#### Assigned result
```{r fig.width=7, fig.height=4}
sa <- sex_annot_bar + mismatch_barplot
sa
```

```{r fig.width=10, fig.height=8}
freq <- sex_assign_out[[3]]
freq
```

```{r}
inconclusive_rows <- which(sex_assign_out[[2]]$assigned_sex == "inconclusive")

kbl(sex_assign_out[[2]], caption = "Assigned results for all samples") %>%
  kable_paper() %>%
  row_spec(row = inconclusive_rows, bold = TRUE, color = "white", background = "#D7261E") %>%
  scroll_box(width = "620px", height = "300px")
```

#### Heatmap of specific X, Y genes
```{r fig.width=10, fig.height=12, eval=FALSE}
hm <- sex_assign_out[[4]]
hm
```
<details>
  <summary>Details</summary>
    ```{r fig.width=10, fig.height=12, echo=FALSE, eval=TRUE}
    hm <- sex_assign_out[[4]]
    hm
    ```
</details>

### Set default sex
inconclusive samples removed
```{r}
se <- sex_assign_out[[1]]
colnames(se@meta.data)[colnames(se@meta.data) == 'sex'] <- 'orig_sex'
colnames(se@meta.data)[colnames(se@meta.data) == 'predicted_sex'] <- 'sex'

se <- subset(se, subset = sex != 'inconclusive')
```

```{r}
se@meta.data %>%
  group_by(sex) %>% #set the condition
  reframe('count' = n_distinct(sample_id), 'sample_id' = toString(unique(sample_id))) %>%
  kbl() %>%
  kable_paper() %>%
  scroll_box(width = "620px", height = "300px")
```

## Ages
```{r fig.width=4, fig.height=4}
if ('project_id' %in% colnames(se@meta.data)) {
  age_distr_sex_proj <- se@meta.data %>%
    ggplot(aes(x = age, y=project_id, fill = factor(sex))) +
    geom_density_ridges(alpha = 0.5, rel_min_height = 0.01) +
    labs(x = "Age", y = "Density",
         title = "Age distribution for Males and Females per project",
         fill = "Sex")
  age_distr_sex_proj
} else {
  age_distr_sex_proj <- se@meta.data %>%
    ggplot(aes(x = age, y=..density.., fill = factor(sex))) +
    geom_density(alpha = 0.5) +
    labs(x = "Age", y = "Density",
         title = "Age distribution for Males and Females",
         fill = "Sex") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  age_distr_sex_proj
}
```

```{r}
if ('project_id' %in% colnames(se@meta.data)) {
  se@meta.data %>%
    group_by(age) %>%
    reframe('project'= toString(unique(project_id)),'size' = n_distinct(sample_id), 'sex' = toString(unique(sex))) %>%
    kbl() %>%
    kable_paper() %>%
    scroll_box(width = "620px", height = "300px")
} else {
  se@meta.data %>%
    group_by(age) %>%
    reframe('size' = n_distinct(sample_id), 'sex' = toString(unique(sex))) %>%
    kbl() %>%
    kable_paper() %>%
    scroll_box(width = "620px", height = "300px")
}
```

## Cell counts
### Minimum cell counts per sample
```{r}
cell_count = 200
```

```{r eval=TRUE}
sample_cell_counts <- table(se@meta.data$sample_id)

# Get the names of samples with more than 200 cells
samples_to_keep <- names(sample_cell_counts[sample_cell_counts >= cell_count])

# Subset the Seurat object to include only cells from samples above 200 cells
se <- subset(se, subset = sample_id %in% samples_to_keep)
```

```{r}
setf <- se@meta.data$sex == "F"
setm <- se@meta.data$sex == "M"

f.cells <- colnames(se)[setf]
m.cells <- colnames(se)[setm]

f.subset <- subset(se, cells = f.cells)
m.subset <- subset(se, cells = m.cells)
```

```{r}
f_ids <- unique(f.subset$sample_id)
m_ids <- unique(m.subset$sample_id)
ids <- c(m_ids, f_ids)
se$sample_id <- factor(se$sample_id, levels = ids)

gradient_palette <- colorRampPalette(c("#0050A0", "grey", "#C64700"))
colors <- gradient_palette(length(unique(se$sample_id)))
```

```{r fig.height=4, fig.width=5}
if ('project_id' %in% colnames(se@meta.data)) {
  ggplot(se@meta.data, aes(x = sample_id, fill = sex)) +
    geom_bar() +
    facet_wrap(~ project_id, scales = "free_x", ncol = 3) +
    geom_hline(yintercept = cell_count, color = "black", linetype = "dashed", size = 1) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x = "sample_id", y = "Count", title = "Cell counts")
} else {
  cc <- ggplot(se@meta.data, aes(x = sample_id, fill = sample_id)) +
    geom_bar(position = "dodge") +
    geom_hline(yintercept = cell_count, color = "black", linetype = "dashed", size = 1) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x = "Sample ID", y = "Count") +
    scale_fill_manual(values = colors) +
    theme_minimal()
}
cc
```

```{r}
kbl(t(table(f.subset@meta.data$sample_id)), caption = "Number of immune cells in female samples") %>%
  kable_paper() %>%
  scroll_box(width = "620px", height = "120px")


kbl(t(table(m.subset@meta.data$sample_id)), caption = "Number of immune cells in male samples") %>%
  kable_paper() %>%
  scroll_box(width = "620px", height = "120px")
```

## Projects (require project_id)
```{r fig.width=8, fig.height=6}
if ('project_id' %in% colnames(se@meta.data)) {
  cell_counts <- se@meta.data %>%
    group_by(project_id) %>%
    reframe(cell_count = n())
  
  sex_counts <- se@meta.data %>%
    group_by(project_id, sex) %>%
    reframe(size = n_distinct(sample_id)) %>%
    pivot_wider(names_from = sex, values_from = size, names_prefix = "size_") %>%
    mutate(size_F = ifelse(is.na(size_F), 0, size_F),
           size_M = ifelse(is.na(size_M), 0, size_M))

  project_data <- sex_counts %>%
    mutate(size = size_F + size_M,
           sex_ratio = round(size_F / (size_F + size_M), 2)) %>%
    arrange(desc(sex_ratio)) %>%
    full_join(cell_counts, by = "project_id")
  
  ggplot(project_data, aes(x = sex_ratio, y = size, color = project_id, size = cell_count)) +
    geom_point() +
    geom_text(aes(label = project_id), vjust = -1.8, size = 2.5) +
    labs(x = "Sex Ratio F/(F+M)", y = "Sample Size (F+M)", color = "Project ID", size = "Cell Count") +
    scale_y_continuous(breaks = seq(0, max(project_data$size), by = 1)) +
    theme_minimal() +
    guides(color = FALSE) +
    xlim(0, 1)
}
```

### Set project criteria
```{r}
if ('project_id' %in% colnames(se@meta.data)) {
  proj_criteria <- project_data$size_F < 3 | project_data$size_M < 3 | project_data$cell_count < 1000
  
  bad_proj <- project_data$project_id[proj_criteria]
  bad_proj_row <- which(proj_criteria)

  kbl(project_data) %>%
    kable_paper() %>%
    row_spec(row = bad_proj_row, bold = TRUE, color = "white", background = "#D7261E") %>%
    scroll_box(width = "620px", height = "300px")
}
```

```{r}
#remove projects in bad_proj
if ('project_id' %in% colnames(se@meta.data)) {
se <- subset(se, subset = project_id %in% bad_proj, invert = TRUE)
}
```

```{r fig.width=8, fig.height=6}
if ('project_id' %in% colnames(se@meta.data)) {
  cell_counts <- se@meta.data %>%
    group_by(project_id) %>%
    reframe(cell_count = n())
  
  sex_counts <- se@meta.data %>%
    group_by(project_id, sex) %>%
    reframe(size = n_distinct(sample_id)) %>%
    pivot_wider(names_from = sex, values_from = size, names_prefix = "size_") %>%
    mutate(size_F = ifelse(is.na(size_F), 0, size_F),
           size_M = ifelse(is.na(size_M), 0, size_M))

  project_data <- sex_counts %>%
    mutate(size = size_F + size_M,
           sex_ratio = round(size_F / (size_F + size_M), 2)) %>%
    arrange(desc(sex_ratio)) %>%
    full_join(cell_counts, by = "project_id")
  
  ggplot(project_data, aes(x = sex_ratio, y = size, color = project_id, size = cell_count)) +
    geom_point() +
    geom_text(aes(label = project_id), vjust = -1.8, size = 2.5) +
    labs(x = "Sex Ratio F/(F+M)", y = "Sample Size (F+M)", color = "Project ID", size = "Cell Count") +
    scale_y_continuous(breaks = seq(0, max(project_data$size), by = 1)) +
    theme_minimal() +
    guides(color = FALSE) +
    xlim(0, 1)
}
```

# Quality control (QC)
## Set your threshold from here
```{r}
nFeature_min = 200
nFeature_max = 5000

nCount_min = 500
nCount_max = 30000

percent.mt_min = 0
percent.mt_max = 10
```

```{r fig.width=12, fig.height=4}
layout(matrix(c(1,2,3), nrow = 1))

nFeature <- se@meta.data[,'nFeature']
hist(nFeature, breaks = 50)
# Add vertical lines
abline(v = mean(nFeature), col = "black", lty = "longdash", lwd = 2)
# 2nd standard deviation
abline(v = mean(nFeature) - 2*sd(nFeature), col = "blue", lty = "longdash", lwd = 2)
abline(v = mean(nFeature) + 2*sd(nFeature), col = "blue", lty = "longdash", lwd = 2)
abline(v = nFeature_min, col = "red", lty = "longdash", lwd = 2)
abline(v = nFeature_max, col = "red", lty = "longdash", lwd = 2)

nCount <- se@meta.data[,'nCount']
hist(nCount, breaks = 50)
# Add vertical lines
abline(v = mean(nCount), col = "black", lty = "longdash", lwd = 2)
# 2nd standard deviation
abline(v = mean(nCount) - 2*sd(nCount), col = "blue", lty = "longdash", lwd = 2)
abline(v = mean(nCount) + 2*sd(nCount), col = "blue", lty = "longdash", lwd = 2)
abline(v = nCount_min, col = "red", lty = "longdash", lwd = 2)
abline(v = nCount_max, col = "red", lty = "longdash", lwd = 2)

percent.mt <- se@meta.data[,'percent.mt']
hist(percent.mt, breaks = 50)
# Add vertical lines
abline(v = mean(percent.mt), col = "black", lty = "longdash", lwd = 2)
# 2nd standard deviation
abline(v = mean(percent.mt) - 2*sd(percent.mt), col = "blue", lty = "longdash", lwd = 2)
abline(v = mean(percent.mt) + 2*sd(percent.mt), col = "blue", lty = "longdash", lwd = 2)
abline(v = percent.mt_min, col = "red", lty = "longdash", lwd = 2)
abline(v = percent.mt_max, col = "red", lty = "longdash", lwd = 2)
```

```{r fig.width=13, fig.height=4}
vl1 <- VlnPlot(se, features = "nFeature", pt.size = 0, group.by = "sex", split.by = "sample_id", cols = colors) +
  geom_hline(aes(yintercept = nFeature_min, color = "threshold"), linetype = "longdash") +
  geom_hline(aes(yintercept = nFeature_max, color = "threshold"), linetype = "longdash") +
  labs(color = "") + # Remove title from color legend
  scale_color_manual(values = c("threshold" = "black"))

vl2 <- VlnPlot(se, features = "nCount", pt.size = 0, group.by = "sex", split.by = "sample_id", cols = colors) +
  geom_hline(aes(yintercept = nCount_min, color = "threshold"), linetype = "longdash") +
  geom_hline(aes(yintercept = nCount_max, color = "threshold"), linetype = "longdash") +
  labs(color = "") + # Remove title from color legend
  scale_color_manual(values = c("threshold" = "black"))

vl3 <- VlnPlot(se, features = "percent.mt", pt.size = 0, group.by = "sex", split.by = "sample_id", cols = colors) +
  geom_hline(aes(yintercept = percent.mt_min, color = "threshold"), linetype = "longdash") +
  geom_hline(aes(yintercept = percent.mt_max, color = "threshold"), linetype = "longdash") +
  labs(color = "") + # Remove title from color legend
  scale_color_manual(values = c("threshold" = "black"))

vl <- wrap_plots(vl1,vl2,vl3) + plot_layout(guides = 'collect')
vl
```

## Examine data before filtering
### Samples in the female group
```{r fig.width=5, fig.height=5, eval=FALSE}
n <- 0
f.sample_id <- unique(f.subset@meta.data$sample_id)
for (i in f.sample_id) {
  n <- n + 1
  simp_sample_id <- paste("female", n, sep = "_")
  
  n.cells <- colnames(f.subset)[f.subset@meta.data$sample_id == i]
  n.subset <- subset(f.subset, cells = n.cells)
  
  #add simplified sample name to metadata
  f.subset@meta.data$simp_sample_id[f.subset@meta.data$sample_id == i] <- simp_sample_id
  se@meta.data$simp_sample_id[se@meta.data$sample_id == i] <- simp_sample_id
  raw@meta.data$simp_sample_id[raw@meta.data$sample_id == i] <- simp_sample_id
  
  Idents(n.subset) <- simp_sample_id
  p1 <- VlnPlot(n.subset, features = c("nFeature", "nCount"), pt.size = 0)
  p2 <- FeatureScatter(n.subset, feature1 = "nCount", feature2 = "nFeature")
  print((p1/p2))
}
```
<details>
  <summary>Details</summary>
    ```{r, fig.width=5, fig.height=5, echo=FALSE, eval=TRUE}
    n <- 0
    f.sample_id <- unique(f.subset@meta.data$sample_id)
    for (i in f.sample_id) {
      n <- n + 1
      simp_sample_id <- paste("female", n, sep = "_")
  
      n.cells <- colnames(f.subset)[f.subset@meta.data$sample_id == i]
      n.subset <- subset(f.subset, cells = n.cells)
  
      #add simplified sample name to metadata
      f.subset@meta.data$simp_sample_id[f.subset@meta.data$sample_id == i] <- simp_sample_id
      se@meta.data$simp_sample_id[se@meta.data$sample_id == i] <- simp_sample_id
  
      Idents(n.subset) <- simp_sample_id
      print((VlnPlot(n.subset, features = c("nFeature", "nCount"), pt.size = 0)/FeatureScatter(n.subset, feature1 = "nCount", feature2 = "nFeature")))
    }
    ```
</details>

### Sample in the male group
```{r fig.width=5, fig.height=5, eval=FALSE}
n <- 0
m.sample_id <- unique(m.subset@meta.data$sample_id)
for (i in m.sample_id) {
  n <- n + 1
  simp_sample_id <- paste("male", n, sep = "_")
  
  n.cells <- colnames(m.subset)[m.subset@meta.data$sample_id == i]
  n.subset <- subset(m.subset, cells = n.cells)
  
  #add simplified sample name to metadata
  m.subset@meta.data$simp_sample_id[m.subset@meta.data$sample_id == i] <- simp_sample_id
  se@meta.data$simp_sample_id[se@meta.data$sample_id == i] <- simp_sample_id
  raw@meta.data$simp_sample_id[raw@meta.data$sample_id == i] <- simp_sample_id
  
  Idents(n.subset) <- simp_sample_id
  p1 <- VlnPlot(n.subset, features = c("nFeature", "nCount"), pt.size = 0)
  p2 <- FeatureScatter(n.subset, feature1 = "nCount", feature2 = "nFeature")
  print((p1/p2))
}
```
<details>
  <summary>Details</summary>
    ```{r, fig.width=5, fig.height=5, echo=FALSE, eval=TRUE}
    n <- 0
    m.sample_id <- unique(m.subset@meta.data$sample_id)
    for (i in m.sample_id) {
      n <- n + 1
      simp_sample_id <- paste("male", n, sep = "_")
      
      n.cells <- colnames(m.subset)[m.subset@meta.data$sample_id == i]
      n.subset <- subset(m.subset, cells = n.cells)
  
      #add simplified sample name to metadata
      m.subset@meta.data$simp_sample_id[m.subset@meta.data$sample_id == i] <- simp_sample_id
      se@meta.data$simp_sample_id[se@meta.data$sample_id == i] <- simp_sample_id
      
      Idents(n.subset) <- simp_sample_id
      p1 <- VlnPlot(n.subset, features = c("nFeature", "nCount"), pt.size = 0)
      p2 <- FeatureScatter(n.subset, feature1 = "nCount", feature2 = "nFeature")
      print((p1/p2))
    }
    ```
</details>


## Filtering
```{r}
set1 <- se$nFeature >= nFeature_min & se$nFeature <= nFeature_max
set2 <- se$nCount >= nCount_min & se$nCount <= nCount_max
set3 <- se$percent.mt >= percent.mt_min & se$percent.mt <= percent.mt_max

keep.cells <- colnames(se)[set1 & set2 & set3]

se.subset <- subset(se, cells = keep.cells)
```

### Examine filtered subset
```{r}
# Create tables for number of cells before and after filtering
before_filtering <- table(se@meta.data$simp_sample_id)
before_filtering <- data.frame(simp_sample_id = names(before_filtering), Before_Filtering = as.vector(before_filtering))

after_filtering <- table(se.subset@meta.data$simp_sample_id)
after_filtering <- data.frame(simp_sample_id = names(after_filtering), After_Filtering = as.vector(after_filtering))

# Merge the tables based on row names
comparison_table <- merge(before_filtering, after_filtering, by = "simp_sample_id", all = TRUE)

# Print the comparison table
print(comparison_table)
```

```{r fig.width=5, fig.height=5, eval=FALSE}
sample_names <- unique(se@meta.data$simp_sample_id)

for (i in sample_names) {
  mycells <- colnames(se.subset)[se.subset@meta.data$simp_sample_id == i]
  mysubset <- subset(se.subset, cells = mycells)
  
  Idents(mysubset) <- i
  p1 <- VlnPlot(mysubset, features = c("nFeature", "nCount"), pt.size = 0)
  p2 <- FeatureScatter(mysubset, feature1 = "nCount", feature2 = "nFeature")
  print((p1/p2))
}
```
<details>
  <summary>Details</summary>
    ```{r, fig.width=5, fig.height=5, echo=FALSE, eval=TRUE}
    sample_names <- unique(se@meta.data$simp_sample_id)

    for (i in sample_names) {
     mycells <- colnames(se.subset)[se.subset@meta.data$simp_sample_id == i]
     mysubset <- subset(se.subset, cells = mycells)
  
     Idents(mysubset) <- i
     p1 <- VlnPlot(mysubset, features = c("nFeature", "nCount"), pt.size = 0)
     p2 <- FeatureScatter(mysubset, feature1 = "nCount", feature2 = "nFeature")
     print((p1/p2))
    }
    ```
</details>

# Patchwork
```{r eval=FALSE}
nFeature_min = 200
nFeature_max = 5000

nCount_min = 500
nCount_max = 30000

percent.mt_min = 0
percent.mt_max = 10

layout <- "
AAAAAAAAAA#CCCCCCCCCCCCCCCCCCCCCCCC
##BBBBBB###DDDDDDDDDDDDDDDDDDDDDDDD
"

wrap_plots(sa, cc, fp, mp) + plot_layout(design = layout) +
  plot_annotation(tag_levels = "A", 

                                 caption = "Figure 1. Sex assignments to donors and quality control for cells.") & 
  theme(plot.caption = ggtext::element_textbox_simple(size = 15))
```

# Save
## Set output file
```{r}
se <- se.subset
output_dir <- "data/processed_data/guilliams_liver/"
date <- format(Sys.Date(), "%Y%m%d")
```

```{r eval=FALSE}
saveRDS(se, file = file.path(output_dir, paste0("standardized_3m3f_", date, ".rds")), compress = FALSE)
save(ct_colors, simp_ct_colors, ct_order, simp_ct_order, file = file.path(output_dir, paste0("cellType_", date, ".RData")))
save(sex_assign_out, file = file.path(output_dir, paste0("sexAssign_output_", date, ".RData")))
```

# Notes
```{r}
sessionInfo()
```
