---
title: Preprocess the Seurat Object
author: Tianze Cao
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
#set to project dir
knitr::opts_knit$set(root.dir = "/home/ti4104ca/Projects/Mine")
knitr::opts_chunk$set(echo = TRUE, results = 'hold', fig.show = 'hold', warning = FALSE)
options(future.globals.maxSize= 10000*1024^2)
```

# Load
## Libraries
```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(ggplot2)
  library(dplyr)
  library(magrittr)
  library(kableExtra)
  library(patchwork)
  load("data/processed_data/guilliams_liver/cellType_20240513.RData")
})
```

## Set input file
```{r}
input <- "data/processed_data/guilliams_liver/standardized_3m3f_20240513.rds"
se <- readRDS(input)
#to fix issue with normalization caused by gene names
sce <- as.SingleCellExperiment(se)
se <- as.Seurat(sce)
```

# Processing standarded data
## Normalization
```{r message=FALSE}
if ("project_id" %in% colnames(se@meta.data)) {
  data.list <- SplitObject(se, split.by = "project_id")
  data.list <- lapply(data.list, function(x) SCTransform(x, verbose = FALSE))
  
  features <- SelectIntegrationFeatures(object.list = data.list, nfeatures = 3000)
  data.list <- PrepSCTIntegration(object.list = data.list, anchor.features = features, verbose = FALSE)
  
  # Integration step for projects
  anchors <- FindIntegrationAnchors(object.list = data.list, normalization.method = "SCT",
                                    anchor.features = features, verbose = FALSE)
  
  data <- IntegrateData(anchorset = anchors, normalization.method = "SCT", verbose = FALSE)
  
  data <- RunPCA(data, verbose = FALSE)
  data <- RunUMAP(data, reduction = "pca", dims = 1:30)
  
  group_var <- "project_id"
} else {
  # Alternative workflow for non-project data
  data <- NormalizeData(se,verbose = F)
  data <- FindVariableFeatures(data,verbose=F)
  data <- ScaleData(data,verbose=F)
  data <- RunPCA(data, features =VariableFeatures(data))
  group_var <- "simp_sample_id"
}
```

### Integration result (skipped for one project)
```{r fig.width=8, fig.height=8}
if ("project_id" %in% colnames(data@meta.data)) {
  p1 <- DimPlot(data, reduction = "pca", dims = c(1,2), group.by = group_var, split.by = "sex")
  p2 <- ElbowPlot(data, reduction = "pca", ndims = 50)

  p1/p2
}
```

```{r fig.width=10, fig.height=12, eval=FALSE}
if ("project_id" %in% colnames(data@meta.data)) {
  VlnPlot(combied, features = paste0("PC_", 1:16), pt.size = 0, group.by = group_var)
}
```
<details>
  <summary>Details</summary>
    ```{r, fig.width=10, fig.height=12, echo=FALSE, eval=TRUE}
    if ("project_id" %in% colnames(data@meta.data)) {
      VlnPlot(data, features = paste0("PC_", 1:16), pt.size = 0, group.by = group_var)
    }
    ```
</details>

```{r fig.width=8, fig.height=16}
if ("project_id" %in% colnames(data@meta.data)) {
  p1 <- DimPlot(data, reduction = "umap", group.by = "cell_type", label = TRUE) + theme(legend.position = "bottom")
  p2 <- DimPlot(data, reduction = "umap", group.by = "simp_cell_type", label = TRUE) + theme(legend.position = "bottom")
  
  p1/p2
}
```

## Dimensionality reduction
### Run PCA
```{r message=FALSE}
data <- RunPCA(data, features =VariableFeatures(data))
```

#### Examine components
```{r fig.width=8, fig.height=8}
p1 <- DimPlot(data, reduction = "pca", dims = c(1,2), group.by = group_var, split.by = "sex")
p2 <- ElbowPlot(data, reduction = "pca", ndims = 30)

p1/p2
```

```{r fig.width=10, fig.height=12, eval=FALSE}
VlnPlot(data, features = paste0("PC_", 1:16), pt.size = 0, group.by = group_var)
```
<details>
  <summary>Details</summary>
    ```{r, fig.width=10, fig.height=12, echo=FALSE, eval=TRUE}
    VlnPlot(data, features = paste0("PC_", 1:16), pt.size = 0, group.by = group_var)
    ```
</details>

### Run harmony
#### Set groups to harmonize
```{r}
groups <- c("sample_id")
theta.use <- 1

data <- harmony::RunHarmony(data, group.by.vars=groups, theta =theta.use,  
                      plot_convergence = TRUE, verbose = FALSE)
```

#### Examine components
```{r fig.width=8, fig.height=8}
if (length(groups) > 0) {
  p1 <- DimPlot(data, reduction = "harmony", group.by = group_var, split.by = "sex")
  p2 <- ElbowPlot(data, reduction = "harmony", ndims = 60)
  
  p1/p2
}
```

test
```{r fig.width=16, fig.height=21, eval=FALSE}
harmony_components <- seq(1, 28, by = 2)
plot_list <- list()

for (i in harmony_components) {
  plot <- DimPlot(data, reduction = "harmony", dims = c(i,(i+1)),
                  group.by = groups, split.by = group_var)
  plot_list[[paste("Harmony", i)]] <- plot
}
cowplot::plot_grid(plotlist=plot_list, ncol = 2)
#VlnPlot(data, features = paste0("harmony_", c(3,4,5,6,9,10,13,14,15,16,19,20)), pt.size = 0, group.by = group_var)
```


## Clustering
### Set reduction, dims
TODO: issue with knitting clustree plot
```{r fig.width=12, fig.height=8, message=FALSE}
reduction <- "harmony"
dims <- c(1:3,7:9,11:13,16:18,20:28)

data <- FindNeighbors(data, reduction = reduction, dims = dims, graph.name = paste0("RNA_snn_harmony_theta", theta.use, ".PC", max(dims)), verbose = FALSE)
for (i in seq(0, 1, 0.1)) {
  data <- FindClusters(data, resolution = i, graph.name = paste0("RNA_snn_harmony_theta", theta.use, ".PC", max(dims)), verbose = FALSE)
}

#has problem to knit
#clustree::clustree(data, prefix = paste0("RNA_snn_harmony_theta", theta.use, ".PC", max(dims), "_res."))
```

### Set resolution
```{r fig.width=8, fig.height=8, message=FALSE}
resolution <- 0.6

data <- FindClusters(data, resolution = resolution, graph.name = paste0("RNA_snn_harmony_theta", theta.use, ".PC", max(dims)), verbose = FALSE)
data <- RunUMAP(data, reducton = reduction, dims = dims, verbose = FALSE)

DimPlot(data, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + theme(legend.position = "bottom")
```

# Cell type annotation
## UMAP plots
```{r}
data$cell_type <- factor(data$cell_type, levels=ct_order)
data$simp_cell_type <- factor(data$simp_cell_type, levels=simp_ct_order)
```

### Include unclassified immune cells (skipped if all are classified)
```{r fig.width=8, fig.height=16, eval=FALSE}
if ("Immune unclassified" %in% data$simp_cell_type){
  p1 <- DimPlot(data, reduction = "umap", group.by = "cell_type", label = TRUE) + theme(legend.position = "bottom")
  p2 <- DimPlot(data, reduction = "umap", group.by = "simp_cell_type", label = TRUE) + theme(legend.position = "bottom")
  
  p1/p2
}
```
<details>
  <summary>Details</summary>
    ```{r, fig.width=8, fig.height=16, echo=FALSE, eval=TRUE}
    if ("Immune unclassified" %in% data$simp_cell_type){
      p1 <- DimPlot(data, reduction = "umap", group.by = "cell_type", label = TRUE) + theme(legend.position = "bottom")
      p2 <- DimPlot(data, reduction = "umap", group.by = "simp_cell_type", label = TRUE) + theme(legend.position = "bottom")
      
      p1/p2
    }
    ```
</details>

### Exclude unclassified immune cells
```{r fig.width=12, fig.height=24}
if ("Immune unclassified" %in% data$simp_cell_type){
  classified <- subset(data, simp_cell_type != "Immune unclassified")
  
  p1 <- DimPlot(classified, reduction = "umap", group.by = "cell_type", label = TRUE) + theme(legend.position = "bottom")
  p2 <- DimPlot(classified, reduction = "umap", group.by = "simp_cell_type", label = TRUE) + theme(legend.position = "bottom")
  
  p1/p2
} else {
  umap <- DimPlot(data, reduction = "umap", group.by = "cell_type", cols = ct_colors, label = TRUE) + theme(legend.position = "bottom") + labs(title = NULL)
  p2 <- DimPlot(data, reduction = "umap", group.by = "simp_cell_type", cols = simp_ct_colors, label = TRUE) + theme(legend.position = "bottom") + labs(title = NULL)
  
  umap/p2
}
```

## Marker genes
### Cluster markers
```{r}
Idents(data) <- "seurat_clusters"
```

```{r message=FALSE}
clusters.markers <- FindAllMarkers(data, only.pos = TRUE)
clusters.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> clusters.top10
```

```{r eval=TRUE}
clusters.top10 <- read.csv("data/processed_data/guilliams_liver/top10markers_Clusters_20240516.csv")
```

```{r}
kbl(clusters.top10) %>%
  kable_paper() %>%
  scroll_box(width = "620px", height = "300px")
```

```{r fig.width=18, fig.height=18}
#downsample due to https://github.com/satijalab/seurat/issues/2724
DoHeatmap(subset(data, downsample = 500), features = clusters.top10$gene) + NoLegend()
```

### Cell type markers
```{r}
Idents(data) <- "cell_type"
```

```{r message=FALSE}
cts.top10 <- data.frame()

for (ct in unique(data$cell_type)) {
  # Replace blank spaces with underscores in the cell type string
  ct_name <- gsub(" ", "_", as.character(ct))
  
  ct.markers <- FindMarkers(data, ident.1 = ct, only.pos = TRUE) %>%
    mutate(cell_type = ct_name) %>%
    mutate(gene = rownames(.))
  
  ct.top10.markers <- ct.markers %>% 
    top_n(n = 10, wt = avg_log2FC)

  #save result to one dataframe cts.top10
  cts.top10 <- bind_rows(cts.top10, ct.top10.markers)
}

cts.top10$cell_type <- gsub("_", " ", as.character(cts.top10$cell_type))
cts.top10$cell_type <- factor(cts.top10$cell_type, levels=ct_order)
cts.top10 <- cts.top10 %>%
  arrange(cell_type)
```

```{r eval=TRUE}
cts.top10 <- read.csv("data/processed_data/guilliams_liver/top10markers_CellTypes_20240516.csv")
```

```{r}
kbl(cts.top10) %>%
  kable_paper() %>%
  scroll_box(width = "620px", height = "300px")
```

```{r fig.width=7, fig.height=20}
#downsample due to https://github.com/satijalab/seurat/issues/2724
mk <- DoHeatmap(subset(data, downsample = 500), features = cts.top10$gene, size = 4, angle = 90) + NoLegend()
mk
```

# Patchwork
```{r eval=FALSE}
nFeature_min = 200
nFeature_max = 5000

nCount_min = 500
nCount_max = 30000

percent.mt_min = 0
percent.mt_max = 10

layout <- "
#####FFFF
EEEEEFFFF
EEEEEFFFF
EEEEEFFFF
EEEEEFFFF
EEEEEFFFF
EEEEEFFFF
EEEEEFFFF
EEEEEFFFF
EEEEEFFFF
EEEEEFFFF
"

wrap_plots(umap, mk) + plot_layout(design = layout) +
  plot_annotation(tag_levels = "A", 

                                 caption = "Figure 2. UMAP visualization and markers of cell types.") & 
  theme(plot.caption = ggtext::element_textbox_simple(size = 15))
```

# Save
## Set output file
```{r}
output_dir <- "data/processed_data/guilliams_liver/"
date <- format(Sys.Date(), "%Y%m%d")
```

```{r eval=FALSE}
saveRDS(data, file = file.path(output_dir, paste0("integrated_3m3f_", date, ".rds")), compress = FALSE)
write.table(clusters.top10, file = file.path(output_dir, paste0("top10markers_Clusters_", date,".csv")), sep = ",", row.names = FALSE)
write.table(cts.top10, file = file.path(output_dir, paste0("top10markers_CellTypes_", date, ".csv")), sep = ",", row.names = FALSE)
```

# Notes
```{r}
sessionInfo()
```
