---
title: Cell Type Frequencies
author: "Tianze Cao"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
#set to project dir
knitr::opts_knit$set(root.dir = "/home/ti4104ca/Projects/Mine")
knitr::opts_chunk$set(echo = TRUE, results = 'hold', fig.show = 'hold', warning = FALSE)
options(future.globals.maxSize= 10000*1024^2)
```

# Load
## Libraries
```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(ggplot2)
  library(dplyr)
  library(magrittr)
  library(kableExtra)
  library(dittoSeq)
  library(speckle)
  library(limma)
  library(patchwork)
})
```

## Set input file
```{r}
input <- "data/processed_data/guilliams_liver/standardized_3m3f_20240513.rds"
se <- readRDS(input)
load("data/processed_data/guilliams_liver/cellType_20240513.RData")
```

# Cell type frequeny plot
```{r}
ditto_order <- c(12,5,13,1,11,10,7,3,4,6,8,2,9)
simp_ditto_order <- c(8,3,10,1,9,4,5,6,2,7)
```

## By sex differences
Age should be counted into biological differences when metadata is sufficient.
### Cell types
```{r fig.width=8, fig.height=8}
bar1 <- dittoBarPlot(se, var = "cell_type", group.by = "sex", color.panel = ct_colors, var.labels.reorder = ditto_order, main = NULL, xlab = "Sex")

#dittoBarPlot(se, var = "cell_type", group.by = "sample_id", xlab = NULL,
 #            split.by = "sex", color.panel = ct_colors, var.labels.reorder = ditto_order) +
  #facet_grid(cols = vars(sex), scale="free", space="free") +
  #theme(legend.position = "bottom")
```

```{r fig.width=8, fig.height=8}
if ('project_id' %in% colnames(se@meta.data)) {
  dittoBarPlot(se, var = "cell_type", group.by = "sample_id", xlab = NULL,
               split.by = c("sex", "project_id"), color.panel = ct_colors, var.labels.reorder = ditto_order) +
    facet_grid(rows = vars(project_id), cols = vars(sex),
               scale="free", space="free") +
    theme(legend.position = "bottom")
} else {
  bar2 <- dittoBarPlot(se, var = "cell_type", group.by = "sample_id", main = NULL, xlab = "Donor ID", ylab = NULL,
               split.by = c("platform", "sex"), color.panel = ct_colors, var.labels.reorder = ditto_order) +
    facet_grid(rows = vars(platform), cols = vars(sex),
               scale="free", space="free")
}
```

```{r fig.width=8, fig.height=6}
bar_plot <- bar1 + bar2 + plot_layout(widths = c(1,2), guides = 'collect')
bar_plot
```

### Simplified cell types
```{r fig.width=8, fig.height=7}
dittoBarPlot(se, var = "simp_cell_type", group.by = "sample_id", xlab = NULL,
             split.by = "sex", color.panel = simp_ct_colors, var.labels.reorder = simp_ditto_order) +
  facet_grid(cols = vars(sex), scale="free", space="free") +
  theme(legend.position = "bottom")
```

```{r fig.width=8, fig.height=8}
if ('project_id' %in% colnames(se@meta.data)) {
  dittoBarPlot(se, var = "simp_cell_type", group.by = "sample_id", xlab = NULL,
               split.by = c("sex", "project_id"), color.panel = simp_ct_colors, var.labels.reorder = simp_ditto_order) +
    facet_grid(rows = vars(project_id), cols = vars(sex),
               scale="free", space="free") +
    theme(legend.position = "bottom")
} else {
  dittoBarPlot(se, var = "simp_cell_type", group.by = "sample", xlab = NULL,
               split.by = c("sex", "sample_id"), color.panel = simp_ct_colors, var.labels.reorder = simp_ditto_order) +
    facet_grid(rows = vars(sex), cols = vars(sample_id),
               scale="free", space="free") +
    theme(legend.position = "bottom")
}
```

## By technique differences
### Cell types
```{r fig.width=8, fig.height=8}
if ('project_id' %in% colnames(se@meta.data)) {
  dittoBarPlot(se, var = "cell_type", group.by = "sample_id", xlab = NULL,
               split.by = "project_id", color.panel = ct_colors, var.labels.reorder = ditto_order) +
    facet_grid(cols = vars(project_id), scale="free", space="free") +
    theme(legend.position = "bottom")
}
```

```{r fig.width=8, fig.height=8}
if ('project_id' %in% colnames(se@meta.data)) {
  dittoBarPlot(se, var = "cell_type", group.by = "sample_id", xlab = NULL,
               split.by = c("platform", "project_id"), color.panel = ct_colors, var.labels.reorder = ditto_order) +
    facet_grid(rows = vars(platform), cols = vars(project_id),
               scale="free", space="free") +
    theme(legend.position = "bottom")
} else {
  dittoBarPlot(se, var = "cell_type", group.by = "sample", xlab = NULL,
               split.by = c("platform", "sample_id"), color.panel = ct_colors, var.labels.reorder = ditto_order) +
    facet_grid(rows = vars(platform), cols = vars(sample_id),
               scale="free", space="free") +
    theme(legend.position = "bottom")
}
```

### Simplified cell types
```{r fig.width=8, fig.height=7}
if ('project_id' %in% colnames(se@meta.data)) {
  dittoBarPlot(se, var = "simp_cell_type", group.by = "sample_id", xlab = NULL,
               split.by = "project_id", color.panel = simp_ct_colors, var.labels.reorder = simp_ditto_order) +
    facet_grid(cols = vars(project_id), scale="free", space="free") +
    theme(legend.position = "bottom")
}
```

```{r fig.width=8, fig.height=7}
if ('project_id' %in% colnames(se@meta.data)) {
  dittoBarPlot(se, var = "simp_cell_type", group.by = "sample_id", xlab = NULL,
               split.by = c("platform", "project_id"), color.panel = simp_ct_colors, var.labels.reorder = simp_ditto_order) +
    facet_grid(rows = vars(platform), cols = vars(project_id),
               scale="free", space="free") +
    theme(legend.position = "bottom")
} else {
  dittoBarPlot(se, var = "simp_cell_type", group.by = "sample", xlab = NULL,
               split.by = c("platform", "sample_id"), color.panel = simp_ct_colors, var.labels.reorder = simp_ditto_order) +
    facet_grid(rows = vars(platform), cols = vars(sample_id),
               scale="free", space="free") +
    theme(legend.position = "bottom")
}
```

# Statistic test
cell frequencies
```{r message=FALSE}
#calculates and transforms cell type proportions
props <- getTransformedProps(clusters = se$cell_type,
                             sample = se$sample_id,
                             transform = "logit")

simp_props <- getTransformedProps(clusters = se$simp_cell_type,
                             sample = se$sample_id,
                             transform = "logit")

freq_df <- props$Proportions %>%
        as.data.frame() %>%
        tidyr::pivot_wider(names_from = "clusters",
                    values_from = "Freq",
                    ) %>%
        tibble::column_to_rownames("sample") %>%
        na.omit()

simp_freq_df <- simp_props$Proportions %>%
        as.data.frame() %>%
        tidyr::pivot_wider(names_from = "clusters",
                    values_from = "Freq",
                    ) %>%
        tibble::column_to_rownames("sample") %>%
        na.omit()
```

set variance info
```{r}
#one dornor has multiple samples, on project only
vari_df <- se@meta.data %>%
  distinct(sample_id, sex, platform) %>%
  tibble::remove_rownames() %>%
  tibble::column_to_rownames("sample_id")

#transform into dummy variable matrix
design <- model.matrix(~ 0 + sex + platform, data = vari_df)
```

## Variance permutation
TODO: add age as linear regression variables
```{r}
#permutational multivariate analysis of variance using distance matrices
adonis <- vegan::adonis2(freq_df ~ sex + platform, data = vari_df, method = "euclidean", permutations = 999)

adonis <- adonis %>%
  filter(!grepl("Total", rownames(.))) %>%
  arrange(SumOfSqs)
```

```{r fig.width=8, fig.height=2}
vari_plot <- adonis %>%
  mutate(row_names = factor(rownames(adonis), levels = unique(rownames(adonis)))) %>%
  ggplot(aes(fill = row_names, y = R2, x = 1)) +
  geom_bar(stat="identity", position = "stack") +
  coord_flip() +
  theme_classic() +
  labs(x="", y="Variance explained", fill="Variables",
       title = "Proportion of cell type frequency variance among donors") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

vari_plot
```

## M/F t-test
```{r}
contr <- makeContrasts(sexM-sexF, levels = design)
```

detailed cell types
```{r}
t_test <- propeller.ttest(props,
                       design,
                       contr,
                       robust = TRUE,
                       trend=FALSE,
                       sort=TRUE
)

t_test$PropRatioLog2 <- log2(t_test$PropRatio)
t_test$M_F_Log2Ratio <- log2(t_test$PropMean.sexM / t_test$PropMean.sexF)
t_test$Sign <- ifelse(t_test$P.Value < 0.001, "***",
                      ifelse(t_test$P.Value < 0.01, "**",
                        ifelse(t_test$P.Value < 0.05, "*", "")))

ct_group <- se@meta.data %>%
  distinct(cell_type, simp_cell_type)
t_test <- t_test %>%
  tibble::rownames_to_column(var = "cell_type") %>%
  full_join(ct_group, by = c("cell_type" = "cell_type"))

kbl(t_test) %>%
  kable_paper() %>%
  scroll_box(width = "620px", height = "300px")
```


simplified cell types
```{r}
simp_t_test <- propeller.ttest(simp_props,
                       design,
                       contr,
                       robust = TRUE,
                       trend=FALSE,
                       sort=TRUE
)

simp_t_test$PropRatioLog2 <- log2(simp_t_test$PropRatio)
simp_t_test$M_F_Log2Ratio <- log2(simp_t_test$PropMean.sexM / simp_t_test$PropMean.sexF)
simp_t_test$Sign <- ifelse(simp_t_test$P.Value < 0.001, "***",
                      ifelse(simp_t_test$P.Value < 0.01, "**",
                        ifelse(simp_t_test$P.Value < 0.05, "*", "")))

simp_t_test <- tibble::rownames_to_column(simp_t_test, var = "cell_type")

kbl(simp_t_test) %>%
  kable_paper() %>%
  scroll_box(width = "620px", height = "300px")
```

```{r}
contr <- makeContrasts(sexM-sexF, levels = design)
```

### Dot plot
```{r fig.width=8, fig.height=6}
dot1 <- t_test %>%
  arrange(desc(M_F_Log2Ratio)) %>%
  mutate(cell_type = factor(cell_type, levels = cell_type),
         simp_cell_type = factor(simp_cell_type, levels = simp_ct_order)) %>%
  # plot
  ggplot(aes(x = M_F_Log2Ratio, y = cell_type, color = simp_cell_type)) +
  geom_vline(xintercept = 0, color = "grey", linewidth = 1.5) +
  # add points
  geom_point(size = 7) +
  facet_grid(rows = vars(factor(simp_cell_type,)),
             scales = "free_y", switch = "y", space = "free",
             labeller = label_wrap_gen(width = 10))+
  theme_bw() +
  theme(strip.text.y = element_text(size = 7), title = element_text(margin = margin(b = 0)))+
  # plot title
  labs(title = NULL, y = NULL,
       color = NULL)  +
  # axis limits
  xlim(c(-3, 3)) +
  #axis title
  xlab("Log2 (Male/Female)") +
  scale_color_manual(values = simp_ct_colors)
dot1
```

```{r fig.width=6, fig.height=8}
# Define the classification of cell types into lymphoid or myeloid
lymphoid_cell_types <- c("NK cell", "Cycling T/NK cell", "T cell", "B cell")

# Mutate cell_type into a new column indicating lymphoid or myeloid
t_test <- t_test %>%
  mutate(cell_type_classification = ifelse(simp_cell_type %in% lymphoid_cell_types, "Lymphoid", "Myeloid")) %>%
  mutate(cell_type_classification = factor(cell_type_classification, levels = c("Lymphoid", "Myeloid"))) %>%
  arrange(desc(M_F_Log2Ratio)) %>%
  mutate(cell_type = factor(cell_type, levels = cell_type),
         simp_cell_type = factor(simp_cell_type, levels = simp_ct_order))

# Plot with facet grid based on cell_type_classification
dot2 <- t_test %>%
  ggplot(aes(x = M_F_Log2Ratio, y = cell_type, color = simp_cell_type)) +
  geom_vline(xintercept = 0, color = "grey", linewidth = 1.5) +
  geom_point(size = 7) +
  facet_grid(rows = vars(cell_type_classification),  # Facet based on cell_type_classification
             scales = "free", space = "free_y") +
  theme_bw() +
  theme(title = element_text(margin = margin(b = 0))) +
  labs(title = NULL, y = NULL,
       color = NULL) +
  xlim(c(-3, 3)) +
  xlab("Log2 (Male/Female)") +
  scale_color_manual(values = simp_ct_colors)
dot2
```

### Box plot
```{r fig.width=3, fig.height=2.6}
df <- merge(vari_df, freq_df, by = "row.names")
simp_df <- merge(vari_df, simp_freq_df, by = "row.names")

box_list <- list()

#set here for simplified cell type or not (df, simp_df)
if ("project_id" %in% se@meta.data) {
  melted_df <- reshape2::melt(df, id.vars = c("Row.names", "sex", "platform", "project_id"))
} else {
  melted_df <- reshape2::melt(df, id.vars = c("Row.names", "sex", "platform"))
}

#set here for simplified cell type or not (t_test, simp_t_test)
names(melted_df)[names(melted_df) == "variable"] <- "cell_type"
melted_df <- left_join(melted_df, t_test[, c("cell_type", "P.Value", "Sign")], by = "cell_type")

for (cell_type in ct_order) {
  subset_df <- melted_df[melted_df$cell_type == cell_type, ]
  
  p <- ggplot(subset_df, aes(x = sex, y = value, fill = sex)) +
    geom_boxplot() +
    geom_jitter(width = 0.2, alpha = 0.5) +
    stat_summary(fun = mean, geom = "point", shape = 23, size = 3, color = "black", fill = "white") +
    stat_summary(fun = median, geom = "point", shape = 21, size = 3, color = "black") +
    labs(x = "Sex", y = "Cell Type Frequency", fill = "Sex") +
    ggtitle(cell_type) +
    theme_minimal() + 
    geom_text(data = subset_df, 
              aes(x = 1.5, y = max(subset_df$value) + 0.1 * diff(range(subset_df$value)), 
                label = paste("P =", round(P.Value, 2), Sign)), 
              vjust = 1, color = "black", size = 4)
  print(p)
  box_list[[cell_type]] <- p
}
```

# Patchwork
```{r eval=FALSE}
layout <- "
AACCC
AACCC
AACCC
AACCC
AACCC
AACCC
BBCCC
"

wrap_plots(bar_plot, vari_plot, dot_plot) + plot_layout(design = layout) +
  plot_annotation(tag_levels = "A", 

                                 caption = "Figure 3. Sex effect on cell type frequency.") & 
  theme(plot.caption = ggtext::element_textbox_simple(size = 15)) 
```

# Notes
```{r}
sessionInfo()
```
