---
title: Gene set enrichment analysis
author: "Tianze Cao"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
#set to project dir
knitr::opts_knit$set(root.dir = "/home/ti4104ca/Projects/Mine")
knitr::opts_chunk$set(echo = TRUE, results = 'hold', fig.show = 'hold', warning = FALSE)
options(future.globals.maxSize= 10000*1024^2)
```

# Load
## Libraries
```{r}
suppressPackageStartupMessages({
  library(readr)
  library(clusterProfiler)
  library(enrichplot)
  library(ggplot2)
  library(org.Hs.eg.db)
  library(UpSetR)
  library(Seurat)
  library(dplyr)
  library(tidyr)
  library(pathview)
  library(msigdbr)
  library(patchwork)
  library(ggupset)
  source("code/scripts/gsea_heatmap.R")
})
```

## Set input file
```{r}
#set cell type from here (cell_type or simp_cell_type)
#load("data/processed_data/guilliams_liver/muscat_output_simpCellType_20240418.RData")
load("data/processed_data/guilliams_liver/muscat_output_allCellType_20240514.RData")
load("data/processed_data/guilliams_liver/cellType_20240513.RData")
#ct_order <- simp_ct_order
#ct_colors <- simp_ct_colors

input <- "data/processed_data/guilliams_liver/integrated_3m3f_20240423.rds"
se <- readRDS(input)
se$simp_sample_id <- factor(se$simp_sample_id, levels = c("male_1","male_2","male_3","female_1","female_2","female_3"))

BTM_plus <- read_tsv("data/raw_data/BTM_plus.tsv")
btm_db <- BTM_plus %>%
  gather(key = "term", value = "gene") %>%
  filter(!is.na(gene))

msig_db <- msigdbr(species = "Homo sapiens", category = "H") %>% 
  distinct(gs_name, entrez_gene) %>%
  as.data.frame()
```

# GSEA
```{r}
cell_types <- unique(muscat_output$celltype_de$celltype_de$de_output_tidy$cluster_id)
x_order <- intersect(ct_order, cell_types)

print(x_order)
```

## Create gene lists
Create gene lists (cluster_list)
```{r}
make_gene_list <- function (cell_type) {
    d <- muscat_output$celltype_de$celltype_de$de_output_tidy %>%
                    filter(cluster_id == !!cell_type , contrast =="M-F" ) %>% #positive higher in males
                    dplyr::select(gene, logFC) #ranked based on logfc
    geneList <- d$logFC
    names(geneList) <- as.character(d$gene)
    geneList <- sort(geneList, decreasing = TRUE)
    return(geneList)
}

cluster_list <- lapply(cell_types, make_gene_list)
names(cluster_list) <- cell_types
```

Create KEGG gene lists (cluster_list_kegg)
```{r message=FALSE}
cluster_list_kegg <- lapply(cluster_list, function(x) {

  name_table <- bitr(names(x), fromType = "SYMBOL", toType = "UNIPROT",
            OrgDb = org.Hs.eg.db) %>% distinct(SYMBOL, .keep_all = TRUE)

  names(x) <- name_table$UNIPROT
  #around 5% genes are failed to map
  x <- x[!is.na(names(x))]
  return(x)
})
```

Create MSig gene lists (cluster_list_msig)
```{r message=FALSE}
cluster_list_msig <- lapply(cluster_list, function(x) {

  name_table <- bitr(names(x), fromType = "SYMBOL", toType = "ENTREZID",
            OrgDb = org.Hs.eg.db) %>% distinct(SYMBOL, .keep_all = TRUE)

  names(x) <- name_table$ENTREZID
  #around 5% genes are failed to map
  x <- x[!is.na(names(x))]
  return(x)
})
```

## GSEA with databases
```{r}
#set cell type from here (cell_type or simp_cell_type)
gsea_go <- readRDS("data/processed_data/guilliams_liver/gseaGO_allCellType_20240514.rds")
#gsea_go <- compareCluster(geneClusters = cluster_list,
 #                    fun = gseGO,
  #                   minGSSize = 10,
   #                  pvalueCutoff = 0.05,
    #                 pAdjustMethod = "BH",
     #                OrgDb = org.Hs.eg.db,
      #               ont = "ALL",
       #              keyType = "SYMBOL")
```

```{r}
#set cell type from here (cell_type or simp_cell_type)
gsea_kegg <- readRDS("data/processed_data/guilliams_liver/gseaKEGG_allCellType_20240514.rds")
#gsea_kegg <- compareCluster(geneClusters = cluster_list_kegg,
 #                    fun = gseKEGG,
  #                   minGSSize = 10,
   #                  pvalueCutoff = 0.05,
    #                 pAdjustMethod = "BH",
     #                organism ="hsa",
      #               keyType = "uniprot")
```

```{r}
#set cell type from here (cell_type or simp_cell_type)
gsea_msig <- readRDS("data/processed_data/guilliams_liver/gseaMSig_allCellType_20240514.rds")
#gsea_msig <- compareCluster(cluster_list_msig, 
 #                          fun = "GSEA", 
  #                         TERM2GENE = msig_db, 
   #                        pvalueCutoff = 0.05, 
    #                       pAdjustMethod = "BH", 
     #                      verbose = FALSE)
```

```{r}
#set cell type from here (cell_type or simp_cell_type)
gsea_btm <- readRDS("data/processed_data/guilliams_liver/gseaBTM_allCellType_20240514.rds")
#gsea_btm <- compareCluster(cluster_list, 
 #                          fun = "GSEA",
  #                         TERM2GENE = btm_db,
   #                        pvalueCutoff= 0.05,
    #                       pAdjustMethod = "BH",
     #                      verbose = FALSE)
```

## Enrichment result
### GO
```{r fig.width=22, fig.height=10}
gsea_go <- pairwise_termsim(gsea_go)
p1<-cnetplot(gsea_go, showCategory = 1, cex_label_category = 2.5, cex_category = 5)
p2<-emapplot(gsea_go, showCategory = 4, cex_category=1.5, cex_label_category = 1.5, cex_line = 0.5, shadowtext = FALSE)
p1+p2
```

#### Dotplot
```{r fig.width=12, fig.height=9}
immune_response <- c(
  "inflammatory response",
  "response to cytokine",
  "response to molecule of bacterial origin",
  "response to bacterium",
  "response to lipopolysaccharide",
  "granulocyte chemotaxis"
)

protein_synthesis <- c(
    "ribosome",
    "ribosomal subunit", 
    "large ribosomal subunit",
    "cytosolic large ribosomal subunit",
    "cytoplasmic translation"
    )

miRNA_metabolism <- c(
  "miRNA metabolic process",
  "regulation of miRNA metabolic process",
  "positive regulation of miRNA metabolic process"
)

adaptive_immune <- c(
  "T cell activation",
  "positive regulation of T cell activation",
  "positive regulation of cell-cell adhesion",
  "positive regulation of leukocyte cell-cell adhesion",
  "MHC class II protein complex",
  "MHC protein complex",
  "peptide antigen binding",
  "antigen binding",
  "lumenal side of endoplasmic reticulum membrane",
  "lumenal side of membrane",
  "clathrin-coated endocytic vesicle membrane",
  "clathrin-coated endocytic vesicle"
)

others <- c(
  "positive regulation of programmed cell death",
  "positive regulation of apoptotic process",
  "detoxification",
  "organic acid binding",
  "carboxylic acid binding"
    )

gsea_go@compareClusterResult$grouping <- ifelse(gsea_go@compareClusterResult$Description %in% immune_response, "Immune Response",
                                           ifelse(gsea_go@compareClusterResult$Description %in% adaptive_immune, "Adaptive Immune",
                                                  ifelse(gsea_go@compareClusterResult$Description %in% protein_synthesis, "Protein Synthesis",
                                                         ifelse(gsea_go@compareClusterResult$Description %in% miRNA_metabolism, "miRNA Metabolism",
                                                                ifelse(gsea_go@compareClusterResult$Description %in% others, "Others", "NA")))))

dotplot <- gsea_go %>%
  filter(p.adjust < 0.05) %>%
  dotplot(color = "NES", size = "setSize", showCategory =4, font.size = 10, label_format =  100) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(factor(grouping, levels = c("Immune Response", "Adaptive Immune", "Protein Synthesis", "miRNA Metabolism", "Others")) ~ .,
             scales = "free_y", switch = "y", space = "free",
             labeller = label_wrap_gen(width = 10)) +
  labs(x = NULL) +
  scale_fill_gradient2(low = "#C64700", high = "#0050A0", midpoint = 0, mid = "white",
							  limits = c(-max(abs(gsea_go@compareClusterResult$NES)), max(abs(gsea_go@compareClusterResult$NES))),
							  breaks = c(max(abs(gsea_go@compareClusterResult$NES)),-max(abs(gsea_go@compareClusterResult$NES))),
							  labels = c("Enriched in Male", "Enriched in Female"))

dotplot
```

#### Upset plot
```{r fig.width=8, fig.height=5}
df <- gsea_go@compareClusterResult %>%
  filter(p.adjust < 0.05)
df_list <- split(df, df$Cluster)

new_df_list <- list()

for (i in seq_along(df_list)) {
  cluster_name <- names(df_list)[i]
  cluster_data <- df_list[[i]]

  new_df_list[[cluster_name]] <- cluster_data[["Description"]]
}

upset(fromList(new_df_list), nsets = length(new_df_list), order.by = "freq", text.scale = 1.3)

#for patchwork
upset <- df %>%
  group_by(Description) %>%
  summarize(Cluster = list(Cluster)) %>%
  ggplot(aes(x = Cluster)) +
    geom_bar() +
    scale_x_upset()
```

### KEGG
```{r fig.width=22, fig.height=10}
gsea_kegg <- pairwise_termsim(gsea_kegg)
p1<-cnetplot(gsea_kegg, showCategory = 3, cex_label_category = 2, cex_category = 5)
p2<-emapplot(gsea_kegg, showCategory = 8, cex_category=1, cex_label_category = 1.5, cex_line = 0.5, shadowtext = FALSE)
p1+p2
```

#### Dotplot
```{r fig.width=10, fig.height=9}
Neurodegenerative <-  c(
  "Oxidative phosphorylation",
  "Parkinson disease",
  "Prion disease",
  "Chemical carcinogenesis - reactive oxygen species",
  "Retrograde endocannabinoid signaling"
  )

Immune_Response <- c(
  "Th17 cell differentiation",
  "Th1 and Th2 cell differentiation",
  "Intestinal immune network for IgA production",
  "Asthma",       
  "Hematopoietic cell lineage",
  "Type I diabetes mellitus",
  "Graft-versus-host disease",
  "Antigen processing and presentation"
)

Inflammatory_Signaling <- c(
  "Complement and coagulation cascades",
  "Toll-like receptor signaling pathway",
  "Lipid and atherosclerosis",
  "NF-kappa B signaling pathway",
  "IL-17 signaling pathway",
  "TNF signaling pathway",
  "MAPK signaling pathway",
  "Cytokine-cytokine receptor interaction"
)

Ohters <- c(
  "Cytoskeleton in muscle cells",
  "Coronavirus disease - COVID-19",
  "Ribosome",
  "Viral protein interaction with cytokine and cytokine receptor",
  "Metabolism of xenobiotics by cytochrome P450",
  "Drug metabolism - cytochrome P450"
)


gsea_kegg@compareClusterResult$grouping <- ifelse(gsea_kegg@compareClusterResult$Description %in% Neurodegenerative, "Neurodegenerative",
                                                ifelse(gsea_kegg@compareClusterResult$Description %in% Immune_Response, "Immune Response",
                                                     ifelse(gsea_kegg@compareClusterResult$Description %in% Inflammatory_Signaling, "Inflammatory Signaling",
                                                          ifelse(gsea_kegg@compareClusterResult$Description %in% Ohters, "Ohters", "NA"))))

dotplot_clusters <- gsea_kegg %>%
  filter(p.adjust < 0.05) %>%
  dotplot(color = "NES", size = "setSize", showCategory =8, font.size = 10, label_format =  100) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(factor(grouping, levels = c("Neurodegenerative", "Immune Response", "Inflammatory Signaling", "Ohters", "NA")) ~ .,
             scales = "free_y", switch = "y", space = "free",
             labeller = label_wrap_gen(width = 10)) +
  labs(x = NULL) +
  scale_fill_gradient2(low = "#C64700", high = "#0050A0", midpoint = 0, mid = "white",
							  limits = c(-max(abs(gsea_kegg@compareClusterResult$NES)), max(abs(gsea_kegg@compareClusterResult$NES))),
							  breaks = c(max(abs(gsea_kegg@compareClusterResult$NES)),-max(abs(gsea_kegg@compareClusterResult$NES))),
							  labels = c("Enriched in Male", "Enriched in Female"))

dotplot_clusters
```

#### Upset plot
```{r fig.width=8, fig.height=6}
df <- filter(gsea_kegg@compareClusterResult, p.adjust < 0.05)
df_list <- split(df, df$Cluster)

new_df_list <- list()

for (i in seq_along(df_list)) {
  cluster_name <- names(df_list)[i]  # Get the cluster name
  cluster_data <- df_list[[i]]       # Get the list of dataframes for this cluster

  new_df_list[[cluster_name]] <- cluster_data[["Description"]]
}

upset(fromList(new_df_list), nsets = length(new_df_list), order.by = "freq", text.scale = 1.3)
```

### MSig
```{r fig.width=18, fig.height=9}
gsea_msig <- pairwise_termsim(gsea_msig)
p1<-cnetplot(gsea_msig, showCategory = 1, cex_label_category = 2, cex_category = 4)
p2<-emapplot(gsea_msig, showCategory = 6, cex_category = 0.3, cex_label_category = 1.5, cex_line = 0.5, shadowtext = FALSE)
p1+p2
```

#### Dotplot
```{r}
immune_inflammatory <- c(
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_IL6_JAK_STAT3_SIGNALING",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_ALLOGRAFT_REJECTION"
)

cellular_stress <- c(
  "HALLMARK_UV_RESPONSE_UP",
  "HALLMARK_HYPOXIA",
  "HALLMARK_OXIDATIVE_PHOSPHORYLATION"
)

transcription_growth <- c(
  "HALLMARK_MYC_TARGETS_V1",
  "HALLMARK_IL2_STAT5_SIGNALING",
  "HALLMARK_KRAS_SIGNALING_UP"
)

metabolic_homeostatic <- c(
  "HALLMARK_HEME_METABOLISM",
  "HALLMARK_ADIPOGENESIS"
)

dna_damage_tumor <- c(
  "HALLMARK_COMPLEMENT",
  "HALLMARK_P53_PATHWAY"
)

gsea_msig@compareClusterResult$grouping <- ifelse(gsea_msig@compareClusterResult$Description %in% immune_inflammatory, "Immune Inflammatory",
                                                ifelse(gsea_msig@compareClusterResult$Description %in% cellular_stress, "Cellular Stress",
                                                     ifelse(gsea_msig@compareClusterResult$Description %in% transcription_growth, "Transcription Growth",
                                                          ifelse(gsea_msig@compareClusterResult$Description %in% metabolic_homeostatic, "Metabolic Homeostatic",
                                                                 ifelse(gsea_msig@compareClusterResult$Description %in% dna_damage_tumor, "Dna Damage Tumor", "NA")))))
```

```{r eval=FALSE}
custom_order <- c(
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_COMPLEMENT",
  "HALLMARK_COAGULATION",
  "HALLMARK_ALLOGRAFT_REJECTION",
  "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_IL2_STAT5_SIGNALING",
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "HALLMARK_MYC_TARGETS_V1",
  "HALLMARK_MITOTIC_SPINDLE",
  "HALLMARK_UV_RESPONSE_DN",
  "HALLMARK_ESTROGEN_RESPONSE_EARLY",
  "HALLMARK_HYPOXIA",
  "HALLMARK_KRAS_SIGNALING_DN",
  "HALLMARK_P53_PATHWAY"
)
filtered_data <- gsea_msig %>%
  mutate(ID = factor(ID, levels = custom_order)) %>%
  arrange(ID)
filtered_data@compareClusterResult$ID <- filtered_data@compareClusterResult$Description
```

```{r fig.width=10, fig.height=6}
dotplot_clusters <- gsea_msig%>%
  filter(p.adjust < 0.05) %>%
  dotplot(color = "NES", size = "setSize", showCategory =4, font.size = 10, label_format =  100) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(factor(grouping, levels = c("Immune Inflammatory", "Cellular Stress", "Transcription Growth", "Metabolic Homeostatic", "Dna Damage Tumor")) ~ .,
             scales = "free_y", switch = "y", space = "free",
             labeller = label_wrap_gen(width = 10)) +
  labs(x = NULL) +
  scale_fill_gradient2(low = "#C64700", high = "#0050A0", midpoint = 0, mid = "white",
							  limits = c(-max(abs(gsea_msig@compareClusterResult$NES)), max(abs(gsea_msig@compareClusterResult$NES))),
							  breaks = c(max(abs(gsea_msig@compareClusterResult$NES)),-max(abs(gsea_msig@compareClusterResult$NES))),
							  labels = c("Enriched in Male", "Enriched in Female"))

dotplot_clusters
```

#### Upset plot
```{r fig.width=8, fig.height=6}
df <- filter(gsea_msig@compareClusterResult, p.adjust < 0.05)
df_list <- split(df, df$Cluster)

new_df_list <- list()

for (i in seq_along(df_list)) {
  cluster_name <- names(df_list)[i]  # Get the cluster name
  cluster_data <- df_list[[i]]       # Get the list of dataframes for this cluster

  new_df_list[[cluster_name]] <- cluster_data[["Description"]]
}

upset(fromList(new_df_list), nsets = length(new_df_list), order.by = "freq", text.scale = 1.3)
```

### BTM
```{r fig.width=18, fig.height=9}
gsea_btm <- pairwise_termsim(gsea_btm)
p1<-cnetplot(gsea_btm, showCategory = 4, cex_label_category = 2, cex_category = 2)
p2<-emapplot(gsea_btm, showCategory = 6, cex_category = 0.3, cex_label_category = 1.5, cex_line = 0.5, shadowtext = FALSE)
p1+p2
```

#### Dotplot
```{r fig.width=10, fig.height=6}
dotplot_clusters <- gsea_btm %>%
  filter(p.adjust < 0.05) %>%
  dotplot(color = "NES", size = "setSize", showCategory =8, font.size = 10, label_format =  100) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  #facet_grid(factor(grouping, levels = c("Neurological Disorders", "Cancer Signaling", "Metabolic Regulation", "Immune Reponse")) ~ .,
   #          scales = "free_y", switch = "y", space = "free",
    #         labeller = label_wrap_gen(width = 10)) +
  labs(x = NULL) +
  scale_fill_gradient2(low = "#C64700", high = "#0050A0", midpoint = 0, mid = "white",
							  limits = c(-max(abs(gsea_btm@compareClusterResult$NES)), max(abs(gsea_btm@compareClusterResult$NES))),
							  breaks = c(max(abs(gsea_btm@compareClusterResult$NES)),-max(abs(gsea_btm@compareClusterResult$NES))),
							  labels = c("Enriched in Male", "Enriched in Female"))

dotplot_clusters
```

#### Upset plot
```{r fig.width=8, fig.height=6}
df <- filter(gsea_btm@compareClusterResult, p.adjust < 0.05)
df_list <- split(df, df$Cluster)

new_df_list <- list()

for (i in seq_along(df_list)) {
  cluster_name <- names(df_list)[i]  # Get the cluster name
  cluster_data <- df_list[[i]]       # Get the list of dataframes for this cluster

  new_df_list[[cluster_name]] <- cluster_data[["Description"]]
}

upset(fromList(new_df_list), nsets = length(new_df_list), order.by = "freq", text.scale = 1.3)
```

### Set category and cell_type
#### KEGG-IL-17 signaling pathway-cDC2s
##### logFC dotplot
```{r fig.width=8, fig.height=3}
#Set category
category <- "IL-17 signaling pathway"

genes_to_plot <- gsea_kegg@compareClusterResult %>%
  filter(Description == category) %>%
  pull(core_enrichment) %>%
  strsplit(split = "/") %>%
  unlist() %>%
  bitr(fromType = "UNIPROT", toType = "SYMBOL", OrgDb = org.Hs.eg.db)

# Filter and subset the gene expression data for the specified contrast (M-F)
filtered_genes <- muscat_output$celltype_de$celltype_de$de_output_tidy %>%
  filter(contrast == "M-F", gene %in% genes_to_plot$SYMBOL) %>%
   mutate(cluster_id = factor(cluster_id, levels = ct_order))

# Create a dot plot of log fold changes (logFC) vs. gene names (gene) colored by cluster_id
ggplot(filtered_genes, aes(x = logFC, y = gene, color = cluster_id)) +
  geom_vline(xintercept = 0, color = "grey", linewidth = 1) +
  geom_point(size = 2) +
  labs(x = "Log Fold Change (M-F)", y = "Gene", color = "Cluster ID") +
  theme_minimal() +
  scale_color_manual(values = ct_colors)
```

##### Heatmap
```{r fig.width=8, fig.height=3}
#set cell type from here (cell_type or simp_cell_type) and category
cell_type <- "cDC2s"
category <- "IL-17 signaling pathway"

genes_to_plot <- gsea_kegg@compareClusterResult %>%
  filter(Description == category) %>%
  pull(core_enrichment) %>%
  strsplit(split = "/") %>%
  unlist() %>%
  bitr(fromType = "UNIPROT", toType = "SYMBOL", OrgDb = org.Hs.eg.db)

se.ct <- subset(se, cells = colnames(se)[se@meta.data$cell_type == cell_type])
Idents(se.ct) <- "simp_sample_id"

DoHeatmap(se.ct, size = 3, features = genes_to_plot$SYMBOL)
```

##### Pathview
```{r}
#set cell type from here (cell_type or simp_cell_type) and kegg_id
cell_type <- "cDC2s"
pathway <- "hsa04657" #IL-17 signaling pathway

gene_list <- cluster_list_kegg[[cell_type]]
a <- bitr_kegg(names(gene_list), fromType = "uniprot", toType="kegg", organism="hsa")
names(gene_list) <- a$kegg
```

TODO: show result in knitted file
```{r eval=FALSE}
date <- format(Sys.Date(), "%Y%m%d")
pathway_obj <- pathview(gene.data  = gene_list,
                     pathway.id = pathway,
                     species    = "hsa",
                     limit      = list(gene=1, cpd=1),
                     out.suffix = paste0(cell_type, "_", date)
)
```

#### GO-MHC class II protein complex-cDCS & T cell activation-mono
##### logFC dotplot
```{r fig.width=8, fig.height=4}
#set category
category <- "inflammatory response"

genes_to_plot <- gsea_go@compareClusterResult %>%
  filter(Description == category) %>%
  pull(core_enrichment) %>%
  strsplit(split = "/") %>%
  unlist()

# Filter and subset the gene expression data for the specified contrast (M-F)
filtered_genes <- muscat_output$celltype_de$celltype_de$de_output_tidy %>%
  filter(contrast == "M-F", gene %in% genes_to_plot) %>%
   mutate(cluster_id = factor(cluster_id, levels = ct_order))

# Create a dot plot of log fold changes (logFC) vs. gene names (gene) colored by cluster_id
ggplot(filtered_genes, aes(x = logFC, y = gene, color = cluster_id)) +
  geom_vline(xintercept = 0, color = "grey", linewidth = 1) +
  geom_point(size = 2) +
  labs(x = "Log Fold Change (M-F)", y = "Gene", color = "Cluster ID") +
  theme_minimal() +
  scale_color_manual(values = ct_colors) + 
  theme(axis.text.y = element_text(size = 0))
```

##### Heatmap
```{r fig.width=8, fig.height=4}
#set cell type from here (cell_type or simp_cell_type) and category
cell_type <- c("cDC1s", "cDC2s")
category <- "MHC class II protein complex"

genes_to_plot <- gsea_go@compareClusterResult %>%
  filter(Description == category) %>%
  pull(core_enrichment) %>%
  strsplit(split = "/") %>%
  unlist()

se.ct <- subset(se, cells = colnames(se)[se@meta.data$cell_type %in% cell_type])
Idents(se.ct) <- "simp_sample_id"

heatmap <- DoHeatmap(subset(se.ct, downsample = 500), size = 3, features = genes_to_plot) + NoLegend()
heatmap
```

```{r fig.width=8, fig.height=7}
#set cell type from here (cell_type or simp_cell_type) and category
cell_type <- "Mono+mono derived cells"
category <- "positive regulation of T cell activation"

genes_to_plot <- gsea_go@compareClusterResult %>%
  filter(Description == category) %>%
  filter(cell_type == cell_type) %>%
  pull(core_enrichment) %>%
  strsplit(split = "/") %>%
  unlist()

se.ct <- subset(se, cells = colnames(se)[se@meta.data$cell_type == cell_type])
Idents(se.ct) <- "simp_sample_id"

DoHeatmap(subset(se.ct, downsample = 500), size = 3, features = genes_to_plot) + NoLegend() + 
    theme(text = element_text(size = 10))
```

#### MsigDB-HALLMARK_TNFA_SIGNALING_VIA_NFKB-Tcells
##### logFC dotplot
```{r fig.width=20, fig.height=5}
#set catogory
category <- "HALLMARK_INFLAMMATORY_RESPONSE"

genes_to_plot <- gsea_msig@compareClusterResult %>%
  filter(Description == category) %>%
  pull(core_enrichment) %>%
  strsplit(split = "/") %>%
  unlist() %>%
  bitr(fromType = "ENTREZID", toType = "SYMBOL", OrgDb = org.Hs.eg.db)

# Filter and subset the gene expression data for the specified contrast (M-F)
filtered_genes <- muscat_output$celltype_de$celltype_de$de_output_tidy %>%
  filter(contrast == "M-F", gene %in% genes_to_plot$SYMBOL) %>%
   mutate(cluster_id = factor(cluster_id, levels = ct_order))

# Create a dot plot of log fold changes (logFC) vs. gene names (gene) colored by cluster_id
logfc <- ggplot(filtered_genes, aes(x = logFC, y = gene, color = cluster_id)) +
  geom_vline(xintercept = 0, color = "grey", linewidth = 1) +
  geom_point(size = 2) +
  labs(x = "Log Fold Change (M-F)", y = "Gene", color = "Cluster ID", title = "Hallmark Genes in Inflammatory Response") +
  theme_minimal() +
  scale_color_manual(values = ct_colors) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) + 
  coord_flip() 
logfc
```

##### Heatmap
```{r fig.width=8, fig.height=12}
#set cell type from here (cell_type or simp_cell_type)
cell_type <- "T cells"
category <- "HALLMARK_TNFA_SIGNALING_VIA_NFKB"

genes_to_plot <- gsea_msig@compareClusterResult %>%
  filter(Description == category) %>%
  filter(cell_type == cell_type) %>%
  arrange(p.adjust) %>%
  pull(core_enrichment) %>%
  strsplit(split = "/") %>%
  unlist() %>%
  bitr(fromType = "ENTREZID", toType = "SYMBOL", OrgDb = org.Hs.eg.db)

se.ct <- subset(se, cells = colnames(se)[se@meta.data$cell_type == cell_type])
Idents(se.ct) <- "simp_sample_id"

DoHeatmap(se.ct, size = 3, features = genes_to_plot$SYMBOL) + NoLegend() + 
  theme(axis.text.y = element_text(size = 7))
```

# Patchwork
```{r eval=FALSE}
wrap_plots(dotplot,logfc,upset,heatmap) + plot_layout(widths = c(1,1), heights = c(5,2)) +
  plot_annotation(tag_levels = "A",
                  caption = "Figure 5. GSEA") & 
  theme(plot.caption = ggtext::element_textbox_simple(size = 15)) 
```

# Save
```{r}
output_dir <- "data/processed_data/guilliams_liver/"
date <- format(Sys.Date(), "%Y%m%d")
```

## Set output file
```{r eval=FALSE}
#set cell type from here (cell_type or simp_cell_type)
saveRDS(gsea_go, paste0(output_dir, "gseaGO_allCellType_", date, ".rds"))
saveRDS(gsea_kegg, paste0(output_dir, "gseaKEGG_allCellType_", date, ".rds"))
saveRDS(gsea_msig, paste0(output_dir, "gseaMSig_allCellType_", date, ".rds"))
saveRDS(gsea_btm, paste0(output_dir, "gseaBTM_allCellType_", date, ".rds"))
```

# Notes
```{r}
sessionInfo()
```