---
title: Sex Differential Expression Analysis
author: "Tianze Cao"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
#set to project dir
knitr::opts_knit$set(root.dir = "/home/ti4104ca/Projects/Mine")
knitr::opts_chunk$set(echo = TRUE, results = 'hold', fig.show = 'hold', warning = FALSE)
options(future.globals.maxSize= 10000*1024^2)
```

# Load
## Libraries
```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(SingleCellExperiment)
  library(dplyr)
  library(ggplot2)
  library(ggraph)
  library(igraph)
  library(tidygraph)
  library(muscatWrapper)
  library(nichenetr)
  library(multinichenetr)
  library(circlize)
  library(kableExtra)
  library(patchwork)
  source("code/scripts/multinichenet_circos.R")
  source("code/scripts/multinichenet_network.R")
})
```

## Set input file
```{r}
input <- "data/processed_data/guilliams_liver/integrated_3m3f_20240423.rds"
se <- readRDS(input)
load("data/processed_data/guilliams_liver/cellType_20240513.RData")
```

## Database
Ligand-receptor network (lr_network) and ligand-target matrix (ligand_target_matrix)
```{r}
organism = "human"
```

```{r}
options(timeout = 120)

if(organism == "human"){
  
  lr_network_all = 
    readRDS(url(
      "https://zenodo.org/record/10229222/files/lr_network_human_allInfo_30112033.rds"
      )) %>% 
    mutate(
      ligand = convert_alias_to_symbols(ligand, organism = organism), 
      receptor = convert_alias_to_symbols(receptor, organism = organism))
  
  lr_network_all = lr_network_all  %>% 
    mutate(ligand = make.names(ligand), receptor = make.names(receptor)) 
  
  lr_network = lr_network_all %>% 
    distinct(ligand, receptor)
  
  ligand_target_matrix = readRDS(url(
    "https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds"
    ))
  
  colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  
  lr_network = lr_network %>% filter(ligand %in% colnames(ligand_target_matrix))
  ligand_target_matrix = ligand_target_matrix[, lr_network$ligand %>% unique()]
  
} else if(organism == "mouse"){
  
  lr_network_all = readRDS(url(
    "https://zenodo.org/record/10229222/files/lr_network_mouse_allInfo_30112033.rds"
    )) %>% 
    mutate(
      ligand = convert_alias_to_symbols(ligand, organism = organism), 
      receptor = convert_alias_to_symbols(receptor, organism = organism))
  
  lr_network_all = lr_network_all  %>% 
    mutate(ligand = make.names(ligand), receptor = make.names(receptor)) 
  lr_network = lr_network_all %>% 
    distinct(ligand, receptor)
  
  ligand_target_matrix = readRDS(url(
    "https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final_mouse.rds"
    ))
  
  colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  
  lr_network = lr_network %>% filter(ligand %in% colnames(ligand_target_matrix))
  ligand_target_matrix = ligand_target_matrix[, lr_network$ligand %>% unique()]
  
}
```

Convert the gene symbols used in the our dataset to make them recognizable in NicheNet 2.0. networks
```{r message=FALSE}
#make cell type names into syntactically valid R names (space->dot)
se$cell_type <- make.names(se$cell_type)
se$simp_cell_type <- make.names(se$simp_cell_type)
ct_order <- make.names(ct_order)
simp_ct_order <- make.names(simp_ct_order)
names(ct_colors)<- make.names(names(ct_colors))
names(simp_ct_colors)<- make.names(names(simp_ct_colors))

sce <- as.SingleCellExperiment(se, assay = "RNA")
sce = alias_to_symbol_SCE(sce, "human") %>% makenames_SCE()
```

# MultiNicheNet
https://github.com/saeyslab/multinichenetr/blob/main/vignettes/pairwise_analysis_MISC.md
## Preparation
### Set covariates from here
project: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE135893
```{r}
sample_id <- "sample_id"
group_id <- "sex"

#set cell type from here
celltype_id <- "cell_type"
#celltype_id <- "simp_cell_type"
#ct_order <- simp_ct_order
#ct_colors <- simp_ct_colors

covariates = NA
batches = "platform"
```

Define sender and receiver cell types
```{r}
#all vs all
cell_types <- c(
  "Plasma.cells",
  "T.cells",
  "B.cells",
  "Neutrophils",
  "Mono.mono.derived.cells",
  "Macrophages",
  "cDC2s",
  "cDC1s",
  "Mig.cDCs",
  "pDCs",
  "Circulating.NK.NKT",
  "Resident.NK",
  "Basophils"
)

senders_oi = cell_types
receivers_oi = cell_types
#some issue when knitting 20240514
#senders_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()
#receivers_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()

#If not all the cell types, subset sce
#sce = sce[, SummarizedExperiment::colData(sce)[,celltype_id] %in% c(senders_oi, receivers_oi)]
```

### Set analysis parameters
```{r}
#samples with inconclusive sex are removed
contrasts_oi <- c("'F-M','M-F'")
contrast_tbl <- tibble(contrast = c("F-M", "M-F"),
                        group = c("F", "M"))

min_cells = 10

min_sample_prop = 0.50
fraction_cutoff = 0.05

#Only when you observe aberrant distributions we recommend redoing the analysis with the empirical p-values instead of the regular p-values. In our experience, this is only required for a limited number of datasets.
empirical_pval = FALSE

logFC_threshold = 0.50
p_val_threshold = 0.05
#if the smallest group >= 20 samples, we typically recommend using p_val_adj = TRUE. When the biological difference between the conditions is very large, we typically recommend increasing the logFC_threshold and/or using p_val_adj = TRUE.
p_val_adj = FALSE

top_n_target = 250
cores_system = 8
# use one core per receiver cell type
n.cores = min(cores_system, union(senders_oi, receivers_oi) %>% length()) 
```

### Set prioritization parameters
"We will set our preference for this dataset as follows - and recommend the user to use the same weights by default if they don’t have specific reasons to change some of the parameters"
```{r eval=FALSE}
prioritizing_weights_DE = c("de_ligand" = 1,
                         "de_receptor" = 1)
prioritizing_weights_activity = c("activity_scaled" = 2)

prioritizing_weights_expression_specificity = c("exprs_ligand" = 2,
                         "exprs_receptor" = 2)

prioritizing_weights_expression_sufficiency = c("frac_exprs_ligand_receptor" = 1)

prioritizing_weights_relative_abundance = c( "abund_sender" = 0,
                         "abund_receiver" = 0)

prioritizing_weights = c(prioritizing_weights_DE, 
                         prioritizing_weights_activity, 
                         prioritizing_weights_expression_specificity,
                         prioritizing_weights_expression_sufficiency, 
                         prioritizing_weights_relative_abundance)
```

```{r eval=TRUE}
scenario = "regular"
ligand_activity_down = FALSE
```

### Run CCC analysis
```{r}
#set cell type from here (cell_type or simp_cell_type)
load("data/processed_data/guilliams_liver/multinichenet_output_allCellType_20240516.RData")

#multinichenet_output = multi_nichenet_analysis(
 # sce = sce, 
  #celltype_id = celltype_id, sample_id = sample_id, group_id = group_id, 
  #batches = batches, covariates = covariates, 
  #lr_network = lr_network, ligand_target_matrix = ligand_target_matrix, 
  #contrasts_oi = contrasts_oi, contrast_tbl = contrast_tbl, 
  #senders_oi = senders_oi, receivers_oi = receivers_oi,
  #min_cells = min_cells, 
  #fraction_cutoff = fraction_cutoff, 
  #min_sample_prop = min_sample_prop,
  #scenario = scenario, 
  #ligand_activity_down = ligand_activity_down,
  #logFC_threshold = logFC_threshold, 
  #p_val_threshold = p_val_threshold, 
  #p_val_adj = p_val_adj, 
  #empirical_pval = empirical_pval, 
  #top_n_target = top_n_target, 
  #n.cores = n.cores, 
  #verbose = TRUE
  #)
```

## Visualization of results
### Circos plot of top-prioritized links
top 50 predictions across all contrasts, senders, and receivers of interest.
```{r fig.width=8, fig.height=8}
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 50, 
  rank_per_group = FALSE
  )

prioritized_tbl_oi = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_all$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_all)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

cell_types = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique())
x_order <- intersect(ct_order, cell_types)

prioritized_tbl_oi$receivernum <- as.numeric(factor(prioritized_tbl_oi$receiver, levels = x_order))
prioritized_tbl_oi$sendernum <- as.numeric(factor(prioritized_tbl_oi$sender, levels = x_order))

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)
colors_receiver = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)

colors_sender = ct_colors
colors_receiver = ct_colors

circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

### Differential L-R pairs
Check the top 50 interactions specific for the Male group
```{r fig.width=16, fig.height=16}
group_oi = "M"

#add senders_oi, receivers_oi when needed
prioritized_tbl_oi_M_50 = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  50, 
  groups_oi = group_oi)

plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_M_50 %>% inner_join(lr_network_all))
plot_oi
```

Check the top 50 interactions specific for the Female group
```{r fig.width=16, fig.height=16}
group_oi = "F"

#add senders_oi, receivers_oi when needed
prioritized_tbl_oi_F_50 = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  50, 
  groups_oi = group_oi)

plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_F_50 %>% inner_join(lr_network_all))
plot_oi
```

### Target gene expression
#### Male-receiver_Mono.mono.derived.cells
```{r}
group_oi = "M"
receiver_oi = "Mono.mono.derived.cells"
```

```{r}
lr_target_prior_cor_filtered = multinichenet_output$lr_target_prior_cor %>%
  inner_join(
    multinichenet_output$ligand_activities_targets_DEgenes$ligand_activities %>% 
      distinct(ligand, target, direction_regulation, contrast)
    ) %>% 
  inner_join(contrast_tbl) %>% filter(group == group_oi, receiver == receiver_oi)

lr_target_prior_cor_filtered_up = lr_target_prior_cor_filtered %>% 
  filter(direction_regulation == "up") %>% 
  filter( (rank_of_target < top_n_target) & (pearson > 0.50 | spearman > 0.50))
lr_target_prior_cor_filtered_down = lr_target_prior_cor_filtered %>% 
  filter(direction_regulation == "down") %>% 
  filter( (rank_of_target < top_n_target) & (pearson < -0.50 | spearman < -0.50)) # downregulation -- negative correlation
lr_target_prior_cor_filtered = bind_rows(
  lr_target_prior_cor_filtered_up, 
  lr_target_prior_cor_filtered_down)
```

```{r fig.width=20, fig.height=6}
prioritized_tbl_oi = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  10, 
  groups_oi = group_oi, 
  receivers_oi = receiver_oi)

lr_target_correlation_plot = make_lr_target_correlation_plot(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi,  
  lr_target_prior_cor_filtered , 
  multinichenet_output$grouping_tbl, 
  multinichenet_output$celltype_info, 
  receiver_oi,
  plot_legend = FALSE)
lr_target_correlation_plot$combined_plot
```

#### Male-receiver_cDC2s 
```{r}
group_oi = "M"
receiver_oi = "cDC2s"
```

```{r}
lr_target_prior_cor_filtered = multinichenet_output$lr_target_prior_cor %>%
  inner_join(
    multinichenet_output$ligand_activities_targets_DEgenes$ligand_activities %>% 
      distinct(ligand, target, direction_regulation, contrast)
    ) %>% 
  inner_join(contrast_tbl) %>% filter(group == group_oi, receiver == receiver_oi)

lr_target_prior_cor_filtered_up = lr_target_prior_cor_filtered %>% 
  filter(direction_regulation == "up") %>% 
  filter( (rank_of_target < top_n_target) & (pearson > 0.50 | spearman > 0.50))
lr_target_prior_cor_filtered_down = lr_target_prior_cor_filtered %>% 
  filter(direction_regulation == "down") %>% 
  filter( (rank_of_target < top_n_target) & (pearson < -0.50 | spearman < -0.50)) # downregulation -- negative correlation
lr_target_prior_cor_filtered = bind_rows(
  lr_target_prior_cor_filtered_up, 
  lr_target_prior_cor_filtered_down)
```

```{r fig.width=20, fig.height=6}
prioritized_tbl_oi = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  10, 
  groups_oi = group_oi, 
  receivers_oi = receiver_oi)

lr_target_correlation_plot = make_lr_target_correlation_plot(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi,  
  lr_target_prior_cor_filtered , 
  multinichenet_output$grouping_tbl, 
  multinichenet_output$celltype_info, 
  receiver_oi,
  plot_legend = FALSE)
lr_target_correlation_plot$combined_plot
```

#### Female-receiver_Macrophages 
```{r}
group_oi = "F"
receiver_oi = "Macrophages"
```

```{r}
lr_target_prior_cor_filtered = multinichenet_output$lr_target_prior_cor %>%
  inner_join(
    multinichenet_output$ligand_activities_targets_DEgenes$ligand_activities %>% 
      distinct(ligand, target, direction_regulation, contrast)
    ) %>% 
  inner_join(contrast_tbl) %>% filter(group == group_oi, receiver == receiver_oi)

lr_target_prior_cor_filtered_up = lr_target_prior_cor_filtered %>% 
  filter(direction_regulation == "up") %>% 
  filter( (rank_of_target < top_n_target) & (pearson > 0.50 | spearman > 0.50))
lr_target_prior_cor_filtered_down = lr_target_prior_cor_filtered %>% 
  filter(direction_regulation == "down") %>% 
  filter( (rank_of_target < top_n_target) & (pearson < -0.50 | spearman < -0.50)) # downregulation -- negative correlation
lr_target_prior_cor_filtered = bind_rows(
  lr_target_prior_cor_filtered_up, 
  lr_target_prior_cor_filtered_down)
```

```{r fig.width=20, fig.height=6}
prioritized_tbl_oi = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  10, 
  groups_oi = group_oi, 
  receivers_oi = receiver_oi)

lr_target_correlation_plot = make_lr_target_correlation_plot(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi,  
  lr_target_prior_cor_filtered , 
  multinichenet_output$grouping_tbl, 
  multinichenet_output$celltype_info, 
  receiver_oi,
  plot_legend = FALSE)
lr_target_correlation_plot$combined_plot
```

#### Female-receiver_cDC1s
```{r}
group_oi = "F"
receiver_oi = "cDC1s"
```

```{r}
lr_target_prior_cor_filtered = multinichenet_output$lr_target_prior_cor %>%
  inner_join(
    multinichenet_output$ligand_activities_targets_DEgenes$ligand_activities %>% 
      distinct(ligand, target, direction_regulation, contrast)
    ) %>% 
  inner_join(contrast_tbl) %>% filter(group == group_oi, receiver == receiver_oi)

lr_target_prior_cor_filtered_up = lr_target_prior_cor_filtered %>% 
  filter(direction_regulation == "up") %>% 
  filter( (rank_of_target < top_n_target) & (pearson > 0.50 | spearman > 0.50))
lr_target_prior_cor_filtered_down = lr_target_prior_cor_filtered %>% 
  filter(direction_regulation == "down") %>% 
  filter( (rank_of_target < top_n_target) & (pearson < -0.50 | spearman < -0.50)) # downregulation -- negative correlation
lr_target_prior_cor_filtered = bind_rows(
  lr_target_prior_cor_filtered_up, 
  lr_target_prior_cor_filtered_down)
```

```{r fig.width=20, fig.height=6}
prioritized_tbl_oi = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  10, 
  groups_oi = group_oi, 
  receivers_oi = receiver_oi)

lr_target_correlation_plot = make_lr_target_correlation_plot(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi,  
  lr_target_prior_cor_filtered , 
  multinichenet_output$grouping_tbl, 
  multinichenet_output$celltype_info, 
  receiver_oi,
  plot_legend = FALSE)
lr_target_correlation_plot$combined_plot
```

### Intercellular regulatory network systems view
links are ligand-target links (= gene regulatory links) and not ligand-receptor protein-protein interactions
```{r}
prioritized_tbl_oi = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  100, 
  rank_per_group = FALSE)

lr_target_prior_cor_filtered = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl$group %>% unique() %>% 
  lapply(function(group_oi){
    lr_target_prior_cor_filtered = multinichenet_output$lr_target_prior_cor %>%
      inner_join(
        multinichenet_output$ligand_activities_targets_DEgenes$ligand_activities %>%
          distinct(ligand, target, direction_regulation, contrast)
        ) %>% 
      inner_join(contrast_tbl) %>% filter(group == group_oi)
    
    lr_target_prior_cor_filtered_up = lr_target_prior_cor_filtered %>% 
      filter(direction_regulation == "up") %>% 
      filter( (rank_of_target < top_n_target) & (pearson > 0.50 | spearman > 0.50))
    
    lr_target_prior_cor_filtered_down = lr_target_prior_cor_filtered %>% 
      filter(direction_regulation == "down") %>% 
      filter( (rank_of_target < top_n_target) & (pearson < -0.50 | spearman < -0.50))
    lr_target_prior_cor_filtered = bind_rows(
      lr_target_prior_cor_filtered_up, 
      lr_target_prior_cor_filtered_down
      )
}) %>% bind_rows()

lr_target_df = lr_target_prior_cor_filtered %>% 
  distinct(group, sender, receiver, ligand, receptor, id, target, direction_regulation)
```

```{r fig.width=22, fig.height=10}
network = infer_intercellular_regulatory_network(lr_target_df, prioritized_tbl_oi)
network_graph = visualize_network(network, colors_sender)
network_graph$plot
```

### Further prioritized wirh network
```{r}
prioritized_tbl_oi_network = prioritized_tbl_oi %>% inner_join(
  network$prioritized_lr_interactions)
```

#### Differential L-R pairs
for male group
```{r fig.width=16, fig.height=12}
group_oi = "M"

prioritized_tbl_oi_M = prioritized_tbl_oi_network %>% filter(group == group_oi)

plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_M %>% inner_join(lr_network_all)
  )
plot_oi
```

for female group
```{r fig.width=16, fig.height=12}
group_oi = "F"

prioritized_tbl_oi_M = prioritized_tbl_oi_network %>% filter(group == group_oi)

plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_M %>% inner_join(lr_network_all)
  )
plot_oi
```

# Save
```{r}
output_dir <- "data/processed_data/guilliams_liver/"
date <- format(Sys.Date(), "%Y%m%d")
```

## Set output file
```{r eval=FALSE}
#set cell type from here (cell_type or simp_cell_type)
save(multinichenet_output, file = file.path(output_dir, paste0("multinichenet_output_allCellType_", date, ".RData")))
```

# Notes
```{r}
sessionInfo()
```
